# Annotations {#annotations}

```{r, include = FALSE}
source("common.R")
columns(1, 2 / 3)
```

## Labels {#sec:labelling}
\index{Labels} \index{Text} \indexf{geom\_text}

Adding text to a plot can be quite tricky. ggplot2 doesn't have all the answers, but does provide some tools to make your life a little easier. The main tool is `geom_text()`, which adds `label`s at the specified `x` and `y` positions. 

`geom_text()` has the most aesthetics of any geom, because there are so many ways to control the appearance of a text:

*   `family` gives the name of a font. There are only three fonts that
    are guaranteed to work everywhere: "sans" (the default), "serif", or "mono":
    
    `r columns(2, 1)`
    ```{r text-family}
    df <- data.frame(x = 1, y = 3:1, family = c("sans", "serif", "mono"))
    ggplot(df, aes(x, y)) + 
      geom_text(aes(label = family, family = family))
    ```
    
    It's trickier to include a system font on a plot because text drawing is 
    done differently by each graphics device (GD). There are five GDs in common 
    use (`png()`, `pdf()`, on screen devices for Windows, Mac and Linux), so
    to have a font work everywhere you need to configure five devices in five
    different ways. Two packages simplify the quandary a bit: 
    
    * showtext, <https://github.com/yixuan/showtext>, by Yixuan Qiu, makes 
      GD-independent plots by rendering all text as polygons. 
    
    * extrafont, <https://github.com/wch/extrafont>, by Winston Chang, 
      converts fonts to a standard format that all devices can use. 
    
    Both approaches have pros and cons, so you will to need to try both of 
    them and see which works best for your needs. \index{Font!family}
    
*   `fontface` specifies the face: "plain" (the default), "bold" or "italic".
    \index{Font!face}

    ```{r text-face}
    df <- data.frame(x = 1, y = 3:1, face = c("plain", "bold", "italic"))
    ggplot(df, aes(x, y)) + 
      geom_text(aes(label = face, fontface = face))
    ```

*   You can adjust the alignment of the text with the `hjust` ("left", "center", 
    "right", "inward", "outward") and `vjust` ("bottom", "middle", "top", 
    "inward", "outward") aesthetics. The default alignment is centered. One
    of the most useful alignments is "inward": it aligns text towards the middle
    of the plot: \index{Font!justification}
    
    ```{r text-justification}
    df <- data.frame(
      x = c(1, 1, 2, 2, 1.5),
      y = c(1, 2, 1, 2, 1.5),
      text = c(
        "bottom-left", "bottom-right", 
        "top-left", "top-right", "center"
      )
    )
    ggplot(df, aes(x, y)) +
      geom_text(aes(label = text))
    ggplot(df, aes(x, y)) +
      geom_text(aes(label = text), vjust = "inward", hjust = "inward")
    ```

*   `size` controls the font size. Unlike most tools, ggplot2 uses mm, rather
    than the usual points (pts). This makes it consistent with other size
    units in ggplot2. (There are 72.27 pts in a inch, so to convert from 
    points to mm, just multiply by 72.27 / 25.4). \index{Font!size}

*   `angle` specifies the rotation of the text in degrees.

You can map data values to these aesthetics, but use restraint: it is hard to percieve the relationship between variables mapped to these aesthetics. `geom_text()` also has three parameters. Unlike the aesthetics, these only take single values, so they must be the same for all labels:

*   Often you want to label existing points on the plot. You don't want the
    text to overlap with the points (or bars etc), so it's useful to offset
    the text a little. The `nudge_x` and `nudge_y` parameters allow you to
    nudge the text a little horizontally or vertically:
    
    ```{r text-nudge}
    df <- data.frame(trt = c("a", "b", "c"), resp = c(1.2, 3.4, 2.5))
    ggplot(df, aes(resp, trt)) + 
      geom_point() + 
      geom_text(aes(label = paste0("(", resp, ")")), nudge_y = -0.25) + 
      xlim(1, 3.6)
    ```
    
    (Note that I manually tweaked the x-axis limits to make sure all the 
    text fit on the plot.)

*   If `check_overlap = TRUE`, overlapping labels will be automatically 
    removed. The algorithm is simple: labels are plotted in the order they 
    appear in the data frame; if a label would overlap with an existing point, 
    it's omitted. This is not incredibly useful, but can be handy. 
    \indexc{check\_overlap}
    
    ```{r text-overlap}
    ggplot(mpg, aes(displ, hwy)) + 
      geom_text(aes(label = model)) + 
      xlim(1, 8)
    ggplot(mpg, aes(displ, hwy)) + 
      geom_text(aes(label = model), check_overlap = TRUE) + 
      xlim(1, 8)
    ```

A variation on `geom_text()` is `geom_label()`: it draws a rounded rectangle behind the text. This makes it useful for adding labels to plots with busy backgrounds: \indexf{geom\_label}

`r columns(1, 2/3)`
```{r label}
label <- data.frame(
  waiting = c(55, 80), 
  eruptions = c(2, 4.3), 
  label = c("peak one", "peak two")
)

ggplot(faithfuld, aes(waiting, eruptions)) +
  geom_tile(aes(fill = density)) + 
  geom_label(data = label, aes(label = label))
```

Labelling data well poses some challenges:

*   Text does not affect the limits of the plot. Unfortunately there's no way
    to make this work since a label has an absolute size (e.g. 3 cm), regardless
    of the size of the plot. This means that the limits of a plot would need to
    be different depending on the size of the plot --- there's just no way to
    make that happen with ggplot2. Instead, you'll need to tweak `xlim()` and
    `ylim()` based on your data and plot size.
    
*   If you want to label many points, it is difficult to avoid overlaps. 
    `check_overlap = TRUE` is useful, but offers little control over which
    labels are removed. There are a number of techniques available for base 
    graphics, like `maptools::pointLabel()`, but they're not trivial to port 
    to the grid graphics used by ggplot2. If all else fails, you may need to
    manually label points in a drawing tool.

Text labels can also serve as an alternative to a legend. This usually makes the plot easier to read because it puts the labels closer to the data. The [directlabels](https://github.com/tdhock/directlabels) package, by Toby Dylan Hocking, provides a number of tools to make this easier: \index{directlabels}

`r columns(2, 1)`
```{r}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()

ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point(show.legend = FALSE) +
  directlabels::geom_dl(aes(label = class), method = "smart.grid")
```

Directlabels provides a number of position methods. `smart.grid` is a reasonable place to start for scatterplots, but there are other methods that are more useful for frequency polygons and line plots. See the directlabels website, <http://directlabels.r-forge.r-project.org>, for other techniques.

## Annotations {#sec:annotations}

Annotations add metadata to your plot. But metadata is just data, so you can use: \index{Annotation} \index{Metadata}

* `geom_text()` to add text descriptions or to label points  Most plots 
  will not benefit from adding text to every single observation on the plot,
  but labelling outliers and other important points is very useful.
  \index{Labels} \indexf{geom\_text}

* `geom_rect()` to highlight interesting rectangular regions of the plot.
  `geom_rect()` has aesthetics `xmin`, `xmax`, `ymin` and `ymax`. 
   \indexf{geom\_rect}

* `geom_line()`, `geom_path()` and `geom_segment()` to add lines.  All 
  these geoms have an `arrow` parameter, which allows you to place an arrowhead 
  on the line.  Create arrowheads with `arrow()`, which has 
  arguments `angle`, `length`, `ends` and `type`. \indexf{geom\_line}

* `geom_vline()`, `geom_hline()` and `geom_abline()` allow you to add
  reference lines (sometimes called rules), that span the full range of the 
  plot. \indexf{geom\_vline} \indexf{geom\_hline} \indexf{geom\_abline}

Typically, you can either put annotations in the foreground (using `alpha` if needed so you can still see the data), or in the background. With the default background, a thick white line makes a useful reference: it's easy to see but it doesn't jump out at you.

To show off the basic idea, we'll draw a time series of unemployment:

`r columns(1, 3 / 8, 1)`
```{r umep}
ggplot(economics, aes(date, unemploy)) + 
  geom_line()
```

We can annotate this plot with which president was in power at the time. There is little new in this code - it's a straightforward manipulation of existing geoms. There is one special thing to note: the use of `-Inf` and `Inf` as positions. These refer to the top and bottom (or left and right) limits of the plot. \indexc{Inf}

```{r unemp-pres}
presidential <- subset(presidential, start > economics$date[1])

ggplot(economics) + 
  geom_rect(
    aes(xmin = start, xmax = end, fill = party), 
    ymin = -Inf, ymax = Inf, alpha = 0.2, 
    data = presidential
  ) + 
  geom_vline(
    aes(xintercept = as.numeric(start)), 
    data = presidential,
    colour = "grey50", alpha = 0.5
  ) + 
  geom_text(
    aes(x = start, y = 2500, label = name), 
    data = presidential, 
    size = 3, vjust = 0, hjust = 0, nudge_x = 50
  ) + 
  geom_line(aes(date, unemploy)) + 
  scale_fill_manual(values = c("blue", "red")) +
  xlab("date") + 
  ylab("unemployment")
```

You can use the same technique to add a single annotation to a plot, but it's a bit fiddly because you have to create a one row data frame:

```{r, fig.show = "hide"}
yrng <- range(economics$unemploy)
xrng <- range(economics$date)
caption <- paste(strwrap("Unemployment rates in the US have 
  varied a lot over the years", 40), collapse = "\n")

ggplot(economics, aes(date, unemploy)) + 
  geom_line() + 
  geom_text(
    aes(x, y, label = caption), 
    data = data.frame(x = xrng[1], y = yrng[2], caption = caption), 
    hjust = 0, vjust = 1, size = 4
  )
```

It's easier to use the `annotate()` helper function which creates the data frame for you: \indexf{annotate}

```{r}
ggplot(economics, aes(date, unemploy)) + 
  geom_line() + 
  annotate("text", x = xrng[1], y = yrng[2], label = caption,
    hjust = 0, vjust = 1, size = 4
  )
```

Annotations, particularly reference lines, are also useful when comparing groups across facets. In the following plot, it's much easier to see the subtle differences if we add a reference line.

`r columns(1, 1 / 4, 1)`
```{r}
ggplot(diamonds, aes(log10(carat), log10(price))) + 
  geom_bin2d() + 
  facet_wrap(~cut, nrow = 1)

mod_coef <- coef(lm(log10(price) ~ log10(carat), data = diamonds))
ggplot(diamonds, aes(log10(carat), log10(price))) + 
  geom_bin2d() + 
  geom_abline(intercept = mod_coef[1], slope = mod_coef[2], 
    colour = "white", size = 1) + 
  facet_wrap(~cut, nrow = 1)
```
