```{r layers, echo=FALSE}
source("common.R")
columns(1, 2 / 3)
```

# Build a plot layer by layer {#cha:layers}

## Introduction

One of the key ideas behind ggplot2 is that it allows you to easily iterate, building up a complex plot a layer at a time. Each layer can come from a different dataset and have a different aesthetic mappings, making it possible to create sophisticated plots that display data from multiple sources. 

You've already created layers with functions like `geom_point()` and `geom_histogram()`. In this chapter, you'll dive into the details of a layer, and how you can control all five components: data, the aesthetic mappings, the geom, stat, and position adjustments. The goal here is to give you the tools to build sophisticated plots tailored to the problem at hand. This more theoretical chapter is accompanied by the next chapter, the "toolbox", which is more hands on, applying the basic components of a layer to specific visualisation challenges.

## Building a plot

So far, whenever we've created a plot with `ggplot()`, we've immediately added on a layer with a geom function. But it's important to realise that there really are two distinct steps. First we create a plot with default dataset and aesthetic mappings:

```{r layer1, error = TRUE}
p <- ggplot(mpg, aes(displ, hwy))
p
```

The plot can't be displayed until we add a layer: there is nothing to see!

```{r}
p + geom_point()
```

`geom_point()` is a shortcut. Behind the scenes it calls the `layer()` function to create a new layer:

```{r layer2, eval = FALSE}
p + layer(
  mapping = NULL, 
  data = NULL,
  geom = "point", geom_params = list(),
  stat = "identity", stat_params = list(),
  position = "identity"
)
```

This call fully specifies the five components to the layer:

*   __mapping__: A set of aesthetic mappings, specified using the 
    `aes()` function and combined with the plot defaults as described in 
    [aesthetic mappings](#sec:aes). If `NULL`, uses the default mapping
    set in `ggplot()`.

*   __data__: A dataset which overrides the default plot dataset. It is usually
    omitted (set to `NULL`), in which case the layer will use the default 
    data specified in `ggplot()`. The requirements for data are explained
    in more detail in [data](#sec:data).
  
*   __geom__: The name of the geometric object to use to draw each observation. 
    Geoms are discussed in more detail in [geom](#sec:data), and 
    [the toolbox](#cha:toolbox) explores their use in more depth.

    Geoms can have additional arguments. All geoms take aesthetics as 
    parameters. If you supply an aesthetic (e.g. colour) as a parameter, 
    it will not be scaled, allowing you to control the appearance of the plot,
    as described in [setting vs. mapping](#sub:setting-mapping). You can
    pass params in `...` (in which case stat and geom parameters
    are automatically teased apart), or in a list passed to `geom_params`.

*   __stat__: The name of the statistical tranformation to use. A statistical
    transformation performs some useful statistical summary is key to histograms 
    and smoothes. To keep the data as is, use the "identity" stat. Learn more 
    in [statistical transformations](#sec:stat).
    
    You only need to set one of stat and geom: every geom has a default stat, 
    and every stat a default geom. 
    
    Most stats take additional parameters to specify the details of statistical 
    transformation. You can supply params either in `...` (in which case stat 
    and geom parameters are automatically teased apart), or in a list called
    `stat_params`.

*   __position__: The method used to adjusting overlapping objects, like
    jittering, stacking or dodging. More details in [position](#sec:position).

It's useful to understand the `layer()` function so you have a better mental model of the layer object. But you'll rarely use the full `layer()` call because it's so verbose. Instead, you'll use the shortcut `geom_` functions: `geom_point(mapping, data, ...)` is exactly equivalent to `layer(mapping, data, geom = "point", ...)`.

## Data {#sec:data}

Every layer must have some data associated with it, and that data must be in a data frame. This is a strong restriction, but there are good reasons for it: \index{Data} \indexc{data.frame}

* Your data is very important, and it's best to be explicit about it.

* A single data frame is also easier to save than a multitude of vectors, which 
  means it's easier to reproduce your results or send your data to someone else.  

* It enforces a clean separation of concerns: ggplot2 turns data frames into
  visualisations. Other packages can make data frames in the right format 
  (learn more about that in [model visualisation](#sub:modelvis)).

The data on each layer doesn't need to be the same, and it's often useful to combine multiple datasets in a single plot. To illustrate that idea I'm going to generate two new datasets related to the mpg dataset. First I'll fit a loess model and generate predictions from it. (This is what `geom_smooth()` does behind the scenes)

```{r loess-pred}
mod <- loess(hwy ~ displ, data = mpg)
grid <- data.frame(displ = seq(min(mpg$displ), max(mpg$displ), length = 50))
grid$hwy <- predict(mod, newdata = grid)

head(grid)
```

Next, I'll isolate observations that are particularly far away from their predicted values:

```{r loess-outlier}
std_resid <- resid(mod) / mod$s
outlier <- subset(mpg, abs(std_resid) > 2)
```

I've generated these datasets because it's common to enhance the display of raw data with a statistical summary and some annotations. With these new datasets, I can improve our initial scatterplot by overlaying a smoothed line, and labelling the outlying points:

```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  geom_line(data = grid, colour = "blue", size = 1.5) + 
  geom_text(data = outlier, aes(label = model))
```

(The text labels aren't particularly easy to read, but you'll learn how to improve those in [polishing]{#cha:polishing}.)

In this example, every layer uses a different dataset. We could define the same plot in another way, omitting the default dataset:

```{r, eval = FALSE}
ggplot(mapping = aes(displ, hwy)) + 
  geom_point(data = mpg) + 
  geom_line(data = grid) + 
  geom_text(data = outlier, aes(label = model))
```

For this case, I don't particularly like this style because it makes it less clear what the primary dataset is (and because of the way that the arguments to `ggplot()` are ordered, it actually requires more keypresses!). However, you may prefer it in cases where there isn't a clear primary dataset, or where the aesthetics also vary from layer to layer. 

NB: if you omit the data set in the call to `ggplot()` you must explicitly supply a dataset for every layer. Also note that facetting will not work without a default dataset: faceting affects all layers so it needs to have a base dataset that defines the set of facets. See [missing faceting variables](#sub:missing-faceting-columns) for more details. 

### Exercises

1.  The first two arguments to ggplot are `data` and `mapping`. The first
    two arguments to all layer functions are `mapping` and `data`. Why does the
    order of the arguments differ? (Hint: think about what you set most 
    commonly.)

1.  The following code uses dplyr to generate some summary statistics about
    each class of car (you'll learn how it works in 
    [data transformation](#cha:dplyr)). 

    ```{r}
    library(dplyr)
    class <- mpg %>% 
      group_by(class) %>% 
      summarise(n = n(), hwy = mean(hwy))
    ```
    
    Use the data to recreate this plot:
    
    ```{r, echo = FALSE}
    ggplot(mpg, aes(class, hwy)) + 
      geom_jitter(width = 0.25) + 
      geom_point(data = class, colour = "red", size = 4) + 
      geom_text(aes(y = 10, label = paste0("n = ", n)), class, size = 3)
    ```

## Aesthetic mappings {#sec:aes}

The aesthetic mappings, defined with `aes()`, describe how variables are mapped to visual properties or __aesthetics__. `aes()` takes a sequence of aesthetic-variable pairs like this: \index{Mappings} \index{Aesthetics!mappings|see{Mappings}} \indexf{aes}

```{r layer11, results = "hide"}
aes(x = displ, y = hwy, colour = class)
```

(If you're American, you can use _color_, and behind the scenes ggplot2 will correct your spelling ;)

Here we map x-position to `displ`, y-position to `hwy`, and colour to `class`. The names for the first two arguments can be ommitted, in which case they correspond to the x and y variables. That makes this specification equivalent to the one above:

```{r layer12, results = "hide"}
aes(displ, hwy, colour = class)
```

While you can do data manipulation in `aes()`, e.g. `aes(log(carat), log(price))`, best to only do simple calculations. It's better to move complex transformations out of `aes()` call and into an explicit `mutate()` call, as you'll learn about in [mutate](#mutate). This makes it easier to check your work and it's often faster (because you need only do the transformation once, not every time the plot is drawn).

Avoid referring to variables that are not in the data (e.g., with `diamonds$carat`). This breaks containment, so that the plot no longer contains everything it needs. This model is a slight simplification: every ggplot has an environment associated with it, so you can refer to objects in that environment and it will work. However, it's best not to rely on this as it prevents the plot from being self-contained. ggplot2 was written before I fully understood non-standard evaluation in R, so it's not as reliable as it could be.

### Specifying the aesthetics in the plot vs. in the layers {#sub:plots-and-layers}

Aesthetic mappings can be supplied in the initial `ggplot()` call, in individual layers, or in some combination of both. All of these calls create the same plot specification:

```{r, eval = FALSE}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point()
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(colour = class))
ggplot(mpg, aes(displ)) + 
  geom_point(aes(y = hwy, colour = class))
ggplot(mpg) + 
  geom_point(aes(displ, hwy, colour = class))
```

You can add, override, or remove mappings:

|Operation |Layer aesthetics    |Result                       |
|:---------|:-------------------|:----------------------------|
|Add       |`aes(colour = cyl)` |`aes(mpg, wt, colour = cyl)` |
|Override  |`aes(y = disp)`     |`aes(mpg, disp)`             |
|Remove    |`aes(y = NULL)`     |`aes(mpg)`                   |

If you only have one layer in the plot, the way you specify aesthetics doesn't make any matter. However, the distinction is important when you start adding additional layers. These two plots are both valid and interesting, but focus on quite different aspects of the data:

`r columns(2, 2 / 3)`
```{r, message = FALSE, warning = FALSE}
ggplot(mpg, aes(displ, hwy, colour = class)) + 
  geom_point() + 
  geom_smooth(se = FALSE)

ggplot(mpg, aes(displ, hwy)) + 
  geom_point(aes(colour = class)) + 
  geom_smooth(se = FALSE)
```

Generally, you want to set up the mappings to illuminate the structure underlying the graphic and minimise typing. It may take some time before the best approach is immediately obvious, so if you've iterated your way to a complex graphic, it may be worthwhile to rewrite it to make the structure more clear.

### Setting vs. mapping {#sub:setting-mapping}

Instead of mapping an aesthetic property to a variable, you can set it to a _single_ value by specifying it in the layer parameters. We **map** an aesthetic to a variable (e.g., `aes(colour = cut)`) or **set**  it to a constant (e.g., `colour = "red"`). \index{Aesthetics!setting|textbf}

The following plots are created with similar code, but have rather different outputs. The second plot **maps** (not sets) the colour to the value 'darkblue'. This effectively creates a new variable containing only the value 'darkblue' and then scales it with a colour scale. Because this value is discrete, the default colour scale uses evenly spaced colours on the colour wheel, and since there is only one value this colour is pinkish. 

`r columns(2, 1)`
```{r layer15}
ggplot(mpg, aes(cty, hwy)) + 
  geom_point(colour = "darkblue") 

ggplot(mpg, aes(cty, hwy)) + 
  geom_point(aes(colour = "darkblue"))
```

A third approach is to map the value, but override the default scale:

```{r}
ggplot(mpg, aes(cty, hwy)) + 
  geom_point(aes(colour = "darkblue")) + 
  scale_colour_identity()
```

This is most useful if you always have a column that already contains colours. You'll learn more about that in [the identity scale](#sub:scale-identity).

It's sometimes useful to map aesthetics to constants. For example, if you want to display multiple layers with varying parameters, you can "name" each layer:

`r columns(1, 2 / 3)`
```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  geom_smooth(aes(colour = "loess"), method = "loess", se = FALSE) + 
  geom_smooth(aes(colour = "lm"), method = "lm", se = FALSE)
```

The default legend title isn't very useful, but you can easily override it as you'll learn in [legends and axes](#sec:guides).

### Exercises

1.  Simplify the following plot specifications:


    ```{r, eval = FALSE}
    ggplot(mpg) + 
      geom_point(aes(mpg$disp, mpg$hwy))
    
    ggplot() + 
     geom_point(mapping = aes(y = hwy, x = cty), data = mpg) +
     geom_smooth(data = mpg, mapping = aes(cty, hwy))
      
    ggplot(diamonds, aes(carat, price)) + 
      geom_point(aes(log(brainwt), log(bodywt)), data = msleep)
    ```

1.  What does the following code do? Does it work? Does it make sense?
    Why/why not?
    
    ```{r, eval = FALSE}
    ggplot(mpg) +
      geom_point(aes(class, cty)) + 
      geom_boxplot(aes(trans, hwy))
    ```

## Geoms {#sec:geom}

Geometric objects, or **geoms** for short, perform the actual rendering of the layer, controlling the type of plot that you create. For example, using a point geom will create a scatterplot, while using a line geom will create a line plot. 

* Graphical primitives:
    * `geom_blank()`:
    * `geom_point()`:
    * `geom_path()`:
    * `geom_ribbon()`:
    * `geom_segment()`:
    * `geom_rect()`:
    * `geom_text()`:
  
* One variable:
    * Discrete: 
        * `geom_bar()`:
    * Continuous
        * `geom_histogram()`:
        * `geom_density()`:
        * `geom_dotplot()`:
        * `geom_freqpoly()`:

* Two variables:
    * Both continuous:
        * `geom_jitter()`:
        * `geom_point()`:
        * `geom_quantile()`:
        * `geom_rug()`:
        * `geom_smooth()`:
        * `geom_text()`:
    * Show distribution:
        * `geom_bin2d():`
        * `geom_density2d()`:
        * `geom_hex()`:
    * Both discrete:
        * `geom_count()`:
        * `geom_jitter()`:
    * One continuous, one discrete:
        * `geom_bar(stat = "identity")`
        * `geom_boxplot()`:
        * `geom_dotplot()`:
        * `geom_violin()`:
    * One time, one continuous
        * `geom_area()`:
        * `geom_line()`:
        * `geom_step()`:
    * Display error:
        * `geom_crossbar()`:
        * `geom_errorbar()`:
        * `geom_linerange()`:
        * `geom_pointrange()`:
    * Spatial
        * `geom_map()`:

* Three variables:
    * `geom_contour()`
    * `geom_tile()`:
    * `geom_raster()`:
  
Each geom has a set of aesthetics that it understands, some of which _must_ be provided. For example, the point geoms requires x and y position, and understands colour, size and shape aesthetics. A bar requires height (`ymax`), and understands width, border colour and fill colour. Each geom lists the aesthetics that can use in its docuementation.

Some geoms differ primarily in the way that they are parameterised.  For example, the tile geom is specified in terms of the location of its centre and its height and width, while the rect geom is parameterised in terms of its top (`ymax`), bottom (`ymin`), left (`xmin`) and right (`right`) positions. \indexf{geom_rect}  Internally, the rect geom is described as a polygon, and its parameters are the locations of the four corners. This is useful for non-Cartesian coordinate systems, as you will learn in [position](#cha:position). \index{Geoms!parameterisation}

Every geom has a default statistic, and every statistic a default geom. For example, the bin statistic defaults to using the bar geom to produce a histogram. These defaults are listed in Table \ref{tbl:geom-aesthetics}.  Overriding these defaults will still produce valid plots, but they may violate graphical conventions. See examples in [combining geoms and stats](#sub:new-plot-types). \index{Geoms!defaults}

## Stat {#sec:stat}

A statistical transformation, or **stat**, transforms the data, typically by summarising it in some manner.  For example, a useful stat is the smoother, which calculates the mean of y, conditional on x, subject to some restriction that ensures smoothness. All currently available stats are listed in Table \ref{tbl:stats}.  To make sense in a graphic context a stat must be location-scale invariant: $\mbox{f}(x + a) = \mbox{f}(x) + a$ and $\mbox{f}(b \cdot x) = b \cdot \mbox{f}(x)$.  This ensures that the transformation stays the same when you change the scales of the plot.

\input{tbls/stats}

\begin{table}
  \begin{center}
  \begin{tabular}{lll}
    \toprule
    Aliased geom & Base geom & Changes in default \\
    \midrule
    area      & ribbon & \verb!aes(min = 0, max = y), position = "stack"!  \\
    density   & area   & \verb!stat = "density"!    \\
    freqpoly  & line   & \verb!stat = "bin"!        \\
    histogram & bar    & \verb!stat = "bin"!        \\
    jitter    & point  & \verb!position = "jitter"! \\
    quantile  & line   & \verb!stat = "quantile"!   \\
    smooth    & ribbon & \verb!stat = "smooth"!     \\
    \bottomrule
  \end{tabular}
  \end{center}
  \caption{Geoms that were created by modifying the defaults of another geom.}
  \label{tbl:aliased-geoms}
\end{table}

### Generated variables

A stat takes a dataset as input and returns a dataset as output, and so a stat can add new variables to the original dataset.  It is possible to map aesthetics to these new variables.  For example, `stat_bin`, the statistic used to make histograms, produces the following variables: \index{Stats!creating new variables} \indexf{stat_bin}

* `count`, the number of observations in each bin
* `density`, the density of observations in each bin (percentage of total / bar width)
* `x`, the centre of the bin

These generated variables can be used instead of the variables present in the original dataset.  For example, the default histogram geom assigns the height of the bars to the number of observations (`count`), but if you'd prefer a more traditional histogram, you can use the density (`density`).   The following example shows a density histogram of `carat` from the diamonds dataset.

```{r hist}
ggplot(diamonds, aes(carat)) + 
  geom_histogram(aes(y = ..density..), binwidth = 0.1)
```

The names of generated variables must be surrounded with `..` when used.  This prevents confusion in case the original dataset includes a variable with the same name as a generated variable, and it makes it clear to any later reader of the code that this variable was generated by a stat.  Each statistic lists the variables that it creates in its documentation. \indexc{..}

## Position adjustments {#sec:position}

Position adjustments apply minor tweaks to the position of elements within a layer.  Table \ref{fig:position} lists all of the position adjustments available within `ggplot`.  Position adjustments are normally used with discrete data.  Continuous data typically doesn't overlap exactly, and when it does (because of high data density) minor adjustments, like jittering, are usually insufficient to fix the problem. \index{Position adjustments} \index{Adjustments!position}  \index{Positioning!position adjustments}

\input{tbls/position}

The different types of adjustment are best illustrated with a bar chart.  Figure \ref{fig:position-bar} shows stacking, filling and dodging.  Stacking puts bars on the same x on top of one another; filling does the same, but normalises height to 1; and dodging places the bars side-by-side.  Dodging is rather similar to faceting, and the advantages and disadvantages of each method are described in [dodging vs. faceting](#sub:dodge-vs-facet). For these operations to work, each bar must have the same width and not overlap with any others.  The identity adjustment (i.e., do nothing) doesn't make much sense for bars, but is shown in Figure \ref{fig:position-identity} along with a line plot of the same data for reference. \index{Dodging} \index{Side-by-side|see{Dodging}} \index{Stacking} \indexf{position_dodge} \indexf{position_fill}  \indexf{position_stack} \indexf{geom_bar}

`r columns(3, 2 / 3)`
```{r position-bar}
dplot <- ggplot(diamonds, aes(clarity, fill = cut)) + 
  theme(legend.position = "none")
dplot + geom_bar(position = "stack")
dplot + geom_bar(position = "fill")
dplot + geom_bar(position = "dodge")
```

The identity positon adjustment is not useful for bars, because each bar obscures the bars behind. A frequency polygon is a better technique here.

`r columns(2, 2 / 3)`
```{r position-identity}
dplot + geom_bar(position = "identity")

ggplot(diamonds, aes(clarity, colour = cut)) + 
  geom_freqpoly(aes(group = cut)) + 
  theme(legend.position = "none")
```

## Pulling it all together {#sec:pull-together}

Once you have become comfortable with combining layers, you will be able to create graphics that are both intricate and useful.  The following examples demonstrate some of the ways to use the capabilities of layers that have been introduced in this chapter.  These are just to get you started. You are limited only by your imagination!

### Combining geoms and stats {#sub:new-plot-types}

By connecting geoms with different statistics, you can easily create new graphics. The code below creates three variations on a histogram. They all use the same statistical transformation underlying a histogram (the bin stat), but use different geoms to display the results: the area geom, the point geom and the tile geom.  \index{Geoms!combining with stats} \index{Stats!combining with geoms}

A number of the geoms available in `ggplot` were derived from other geoms in a process like the one just described, starting with an existing geom and making a few changes in the default aesthetics or stat.  For example, the jitter geom is simply the point geom with the default position adjustment set to jitter.  Once it becomes clear that a particular variant is going to be used a lot or used in a very different context, it makes sense to create a new geom. Table \ref{tbl:aliased-geoms} lists these ``aliased'' geoms. \index{Geoms!aliases}

### Varying aesthetics and data {#sub:different-aesthetics}

One of the more powerful capabilities of `ggplot` is the ability to plot different datasets on different layers. This may seem strange: Why would you want to plot different data on the same plot?  In practice, you often have related datasets that should be shown together.  A very common example is supplementing the data with predictions from a model.  While the smooth geom can add a wide range of different smooths to your plot, it is no substitute for an external quantitative model that summarises your understanding of the data.
