---
output: bookdown::html_chapter
bibliography: references.bib
---

```{r polishing, include = FALSE}
source("common.R")
```

# Polishing your plots for publication {#cha:polishing}

In this chapter you will learn how to prepare polished plots for publication.  Most of this chapter focusses on the theming capability of ggplot which allows you to control many non-data aspects of plot appearances, but you'll also learn functions that make adding annotations easier and the best way to save plots for inclusion into other software packages. \index{Publication!polishing plots for}

The visual appearance of the plot is determined by both data and non-data related components.  [Themes](#sec:themes) introduces the theme system which controls all aspects of non-data display.  By now you should be familiar with the many ways that you can alter the data-related components of the plot---layers and scales---to visualise your data and change the appearance of the plot.

[Saving your output](#sec:saving) discusses the chapter with a discussion about how to get your graphics out of R and into LaTeX, Word or other presentation or word-processing software.  [Multiple plots on the same page](#sec:grid-layout) concludes with a discussion of how to lay out multiple plots on a single page.

<!--
% Need cross-refs here to changing axis, legend and plot titles
% And guide breaks and legends
-->

## Themes {#sec:themes}

The appearance of non-data elements of the plot is controlled by the theme system. The theme system does not affect how the data is rendered by geoms, or how it is transformed by scales. Themes don't change the perceptual properties of the plot, but they do help you make the plot aesthetically pleasing or match existing style guides. Themes give you control over things like the fonts in all parts of the plot: the title, axis labels, axis tick labels, strips, legend labels and legend key labels; and the colour of ticks, grid lines and backgrounds (panel, plot, strip and legend). \index{Themes} \index{Publication!themes}

This separation of control into data and non-data parts is quite different than base and lattice graphics.  In base and lattice graphics, most functions take a large number of arguments that specify both data and non-data appearance, which makes the functions complicated and hard to learn. ggplot2 takes a different approach: when creating the plot you determine how the data is displayed, then *after* it has been created you can edit every detail of the rendering, using the theming system.  Some of the effects of changing the theme of a plot are shown in below:

`r columns(3, 1)`
```{r themes}
base <- ggplot(mpg, aes(displ, hwy)) + geom_point()
base
base + theme_bw()
base + theme_minimal()
```

Like many other areas of ggplot2, themes can be controlled on multiple levels from coarse to fine. You can:

* Use a complete theme. This affects every element of the plot in a visually 
  consistent manner.  The default theme uses a grey panel background with white 
  gridlines; however, there are built-in alternatives such as `theme_bw()` 
  which uses a white background with grey gridlines 
  ([link to section](#sec:built-in)).
  
* Modify a single element of a built-in theme. Each theme is made up of 
  multiple elements. The theme system comes with a number of built-in element 
  rendering functions with a limited set of parameters. By adjusting these 
  parameters you can control things like text size and colour, background and 
  grid line colours and text orientation.  By combining multiple elements you 
  can create your own theme ([link to section](#sec:theme-elements)).

### Complete themes {#sec:built-in}

ggplot2 comes with a number of built in themes: \index{Themes!built-in}

* `theme_grey()`: the signature ggplot2 theme with a light grey background and 
  white gridlines, designed to put the data forward yet make comparisons easy.
  This follows from the advice of [@tufte:2006; @tufte:1990; @tufte:2001; 
  @tufte:1997] and [@brewer:1994; @carr:2002; @carr:1994; @carr:1999]. We can 
  still see the gridlines to aid in the judgement of position 
  [@cleveland:1993a], but they have little visual impact and we can easily 
  'tune' them out. The grey background gives the plot a similar typographic 
  colour to the text, ensuring that the graphics fit in with the flow of a 
  document without jumping out with a bright white background. Finally, the 
  grey background creates a continuous field of colour which ensures that the 
  plot is perceived as a single visual entity. \indexf{theme_grey}
  
* `theme_bw()`: the classic dark-on-light ggplot2 theme. May work better for 
  presentations displayed with a projector.
  
* `theme_linedraw()`: A theme with only black lines of various widths on white 
  backgrounds, reminiscent of a line drawing.
  
* `theme_light()`: similar to `theme_linedraw()` but with light grey lines and 
  axes, to direct more attention towards the data.}

* `theme_dark()`: the dark cousin of `theme_light()`, with similar line sizes 
  but a dark background. Useful to make thin coloured lines pop out.
  
* `theme_minimal()`: A minimalistic theme with no background annotations.

* `theme_classic()`: A classic-looking theme, with x and y axis lines and no 
   gridlines.
   
* `theme_void()`: A completely empty theme. 

```{r}
df <- data.frame(x = 1:3, y = 1:3)
base <- ggplot(df, aes(x, y)) + geom_point()
base + theme_grey() 
base + theme_bw()
base + theme_linedraw()
base + theme_light()
base + theme_dark()
base + theme_minimal() 
base + theme_classic()
base + theme_void()
```

All themes have a `base_size` parameter which controls the base font size. The base font size is the size that the axis titles use: the plot title is usually bigger, and the tick and strip labels are smaller. If you want to control these sizes separately, you'll need to modify the individual elements as described in the following section.

You're not limited to just built-in themes. Other packages, like ggtheme, add even more. Here's a selection:

`r columns(3, 1)`
```{r}
library(ggthemes)
base + theme_tufte() 
base + theme_solarized()
base + theme_excel() # ;)
```

### Theme elements and element functions {#sec:theme-elements}

A theme is made up of multiple *elements* which control the appearance of a single item on the plot, as listed in Table~\ref{tbl:elements}. There are three elements that have individual `x` and `y` settings: `axis.text`, `axis.title` and `strip.text`. Having a different setting for the horizontal and vertical elements allows you to control how text should appear in different orientations. The appearance of each element is controlled by an *element function*. \index{Themes!elements}

There are four basic types of built-in element functions: text, lines and segments, rectangles and blank.  Each element function has a set of parameters that control the appearance as described below:

*   `element_text()` draws labels and headings. You can control the font 
    `family`, `face`, `colour`, `size`, `hjust`, `vjust`, `angle` and 
    `lineheight`. \index{Themes!labels} \indexf{element_text}
  
    ```{r theme-text}
    base_t <- base + labs(title = "This is a ggplot")
    base_t
    base_t + theme(plot.title = element_text(size = 20))
    base_t + theme(plot.title = element_text(face = "bold", colour = "red"))
    
    base + theme(axis.text = element_text(colour = "red"))
    base + theme(axis.text.y = element_text(angle = -45))
    base + theme(axis.text.x = element_text(angle = -90, vjust = 0.5))
    ```

*  `element_line()` draws lines parameterised by `colour`, `size` and
   `linetype`: \indexf{element_line}

    ```{r theme-line}
    base + theme(panel.grid.major = element_line(colour = "red"))
    base + theme(panel.grid.major = element_line(size = 2))
    base + theme(panel.grid.major = element_line(linetype = "dotted"))
    base + theme(axis.line = element_line())
    base + theme(axis.line = element_line(colour = "red"))
    base + theme(axis.line = element_line(size = 0.5, linetype = "dashed"))
    ```

* `element_rect()` draws rectangles, mostly used for backgrounds, parameterised
   by `fill` colour and border `colour`, `size` and `linetype`.  
   \index{Background} \index{Themes!background} \indexf{theme_rect}
  
    ```{r theme-background}
    base + theme(plot.background = element_rect(fill = "grey80", colour = NA))
    base + theme(plot.background = element_rect(colour = "red", size = 2))
    base + theme(panel.background = element_rect(fill = "linen"))
    ```

*   `element_blank()` draws nothing.  Use this element if you don't want 
    anything drawn, and no space allocated for that element.  The following
    example uses `element_blank()` to progressively suppress the appearance of 
    elements we're not interested in. Notice how the plot automatically reclaims 
    the space previously used by these elements: if you don't want this to 
    happen (perhaps because they need to line up with other plots on the page), 
    use `colour = NA, fill = NA` as parameter to create invisible elements that 
    still take up space. \indexf{element_blank}

    ```{r theme-blank}
    base
    last_plot() + theme(panel.grid.minor = element_blank())
    last_plot() + theme(panel.grid.major = element_blank())
    last_plot() + theme(panel.background = element_blank())
    last_plot() + theme(
      axis.title.x = element_blank(), 
      axis.title.y = element_blank()
    )
    last_plot() + theme(axis.line = element_line(colour = "grey50"))
    ```

### Theme inheritance

### Global settings

As well as applying themes to each plot individually, you can also set them globally with `theme_set()`.  `theme_set()` returns the previous theme so that you can restore it later if you want. \indexf{theme_set}

You can modify the elements locally for a single plot with `theme()` (as seen above), or globally for all future plots with `theme_update()`. \index{Themes!updating} \indexf{theme_get} \indexf{theme}

```{r theme-update}
old_theme <- theme_update(
  plot.background = element_rect(fill = "lightblue3", colour = NA),
  panel.background = element_rect(fill = "lightblue", colour = NA),
  axis.text = element_text(colour = "#CCFF33"),
  axis.title = element_text(colour = "#CCFF33")
)
base
theme_set(theme_grey())
```

### Exercises

1.  Create the ugliest plot possible! (Contributed by Andrew D. Steen, 
    University of Tennessee - Knoxville)
    
## Annotations {#sec:annotating}

When annotating your plot with additional labels, the important thing to remember is that these annotations are just extra data. You can use \index{Annotation}

* `geom_text()` to add text descriptions or to label points.  Most plots 
  will not benefit from adding text to every single observation on the plot.
  However, pulling out just a few observations (using subset) can be very useful.
  Typically you will want to label outliers or other important points. \index{Labels}

* `geom_rect()` to highlight interesting rectangular regions of the plot.
  `geom_rect()` has aesthetics `xmin`, `xmax`, `ymin` and `ymax`. 
   \indexf{geom_rect}

* `geom_line()`, `geom_path()` and `geom_segment()` to add lines.  All 
  these geoms have an `arrow` parameter, which allows you to place an arrowhead 
  on the line.  Create arrowheads with `grid::arrow()`, which has 
  arguments `angle`, `length`, `ends` and `type`. \indexf{geom_line}

* `geom_vline()`, `geom_hline()` and `geom_abline()` allow you to add
  reference lines (sometimes called rules), that span the full range of the 
  plot. \indexf{geom_vline} \indexf{geom_hline} \indexf{geom_abline}

Typically, you can either put annotations in the foreground (using `alpha` if needed so you can still see the data), or in the background. With the default background, a thick white line makes a useful reference: it's easy to see but it doesn't jump out at you.

To show off the basic idea, we'll draw a time series of unemployment:

`r columns(1, 3 / 8, 1)`
```{r umep}
ggplot(economics, aes(date, unemploy)) + 
  geom_line()
```

We can annotate this plot with which president was in power at the time. There is little new in this code - it's a straightforward manipulation of existing geoms. There is one special thing to note: the use of `-Inf` and `Inf` as positions. These refer to the top and bottom (or left and right) limits of the plot.

```{r unemp-pres}
presidential <- subset(presidential, start > economics$date[1])

ggplot(economics) + 
  geom_rect(
    aes(xmin = start, xmax = end, fill = party), 
    ymin = -Inf, ymax = Inf, alpha = 0.2, 
    data = presidential
  ) + 
  geom_vline(
    aes(xintercept = as.numeric(start)), 
    data = presidential,
    colour = "grey50", alpha = 0.5
  ) + 
  geom_text(
    aes(x = start, y = 2500, label = name), 
    data = presidential, 
    size = 3, vjust = 0, hjust = 0, nudge_x = 50
  ) + 
  geom_line(aes(date, unemploy)) + 
  scale_fill_manual(values = c("blue", "red"))
```

You can use the same technique to add a single annotation to a plot, but it's a bit fiddly because you have to create a one row data frame:

```{r, fig.show="hide"}
yrng <- range(economics$unemploy)
xrng <- range(economics$date)
caption <- paste(strwrap("Unemployment rates in the US have 
  varied a lot over the years", 40), collapse = "\n")

ggplot(economics, aes(date, unemploy)) + 
  geom_line() + 
  geom_text(
    aes(x, y, label = caption), 
    data = data.frame(x = xrng[1], y = yrng[2], caption = caption), 
    hjust = 0, vjust = 1, size = 4
  )
```

It's easier to use the `annotate()` helper function which creates the data frame for you:

```{r}
ggplot(economics, aes(date, unemploy)) + 
  geom_line() + 
  annotate("text", x = xrng[1], y = yrng[2], label = caption,
    hjust = 0, vjust = 1, size = 4
  )
```

Annotations, particularly reference lines are also useful when comparing groups across facets. In the following plot, it's much easier to see the subtle differences if we add a reference line.

`r columns(1, 1 / 4, 1)`
```{r}
ggplot(diamonds, aes(log10(carat), log10(price))) + 
  geom_bin2d() + 
  facet_wrap(~cut, nrow = 1)

mod_coef <- coef(lm(log10(price) ~ log10(carat), data = diamonds))
ggplot(diamonds, aes(log10(carat), log10(price))) + 
  geom_bin2d() + 
  geom_abline(intercept = mod_coef[1], slope = mod_coef[2], 
    colour = "white", size = 1) + 
  facet_wrap(~cut, nrow = 1)
```

## Saving your output {#sec:saving}

You have two basic choices of output: raster or vector.  Vector graphics are procedural.  This means that they are essentially 'infinitely' zoomable; there is no loss of detail.  Raster graphics are stored as an array of pixels and have a fixed optimal viewing size.  Figure \ref{fig:vector-raster} illustrates the basic differences for a basic circle.  A good description is available at <http://tinyurl.com/rstrvctr>. \index{Saving} \index{Exporting} \index{Publication!saving output}

\begin{figure}[htbp]
  \centering
    \includegraphics[width= 0.5\linewidth]{diagrams/vector-raster}
  \caption{The schematic difference between raster (left) and vector (right) graphics. }
  \label{fig:vector-raster}
\end{figure}

Generally, vector output is more desirable, but for complex graphics containing thousands of graphical objects it can be slow to render.  In this case, it may be better to switch to raster output.  For printed use, a high-resolution (e.g., 600 dpi) graphic may be an acceptable compromise, but may be large. 

To save your output, you can use the typical R way with disk-based graphics devices, which works for all packages, or a special function from `ggplot` that saves the current plot: `ggsave()`. `ggsave()` is optimised for interactive use and has the following important arguments: \indexf{ggsave}

* The `path` specifies the path where the image should be saved. The file 
  extension will be used to automatically select the correct graphics device.

* Three arguments control output size.  If left blank, the size of the current 
  on-screen graphics device will be used. Otherwise, override with `width` and 
  `height` arguments. 
  
* For raster graphics, the `dpi` argument controls the resolution of the plot. 
  It defaults to 300, which is appropriate for most printers, but you may want 
  to use 600 for particularly high-resolution output, or 72 for on-screen 
  (e.g., web) display.

The following code shows these two methods. If you want to save multiple plots to a single file, you will need to explicitly open a disk-based graphics device (like `png()` or `pdf()`), print the plots and then close it with `dev.off()`.

```{r ggsave, eval = FALSE}
ggplot(mpg, aes(displ, hwy)) + geom_point()
ggsave("output.pdf")

pdf("output.pdf", width = 6, height = 6)
# If inside a script, you will need to explicitly print() plots
ggplot(mpg, aes(displ, hwy)) + geom_point()
ggplot(mpg, aes(displ, cty)) + geom_point()
dev.off()
```

Table~\ref{tbl:graphic-recommendation} lists recommended graphic formats for various tasks. R output generally works best as part of a linux development tool chain: using png or pdf output in LaTeX documents.  With Microsoft Office it is easiest to use a high-resolution (`dpi = 600`) png file.  You can use vector output, but neither Windows meta files nor postscript supports transparency, and while postscript prints fine, it's not displayed on screen (unless you add a preview image with another tool). Transparency is used to show confidence intervals with the points showing through. If you copy and paste a graph into Word, and see that the confidence interval bands have vanished, that is the cause. The same advice holds for OpenOffice. \index{Exporting!to Word} \index{Exporting!to Powerpoint}

If you are using LaTeX, I recommend including `\DeclareGraphicsExtensions{.png,.pdf}` in the preamble.  Then you don't need to specify the file extension in `\includegraphics{}` commands, but LaTeX will pick png files in preference to pdf. \index{Exporting!to Latex} I choose this order because you can produce all your files in pdf, and then go back and re-render any big ones as png.  Another useful command is `\graphicspath{}` which specifies a path in which to look for graphics, allowing you to keep graphics in a separate directory to the text.

\begin{table}
  \begin{center}
  \begin{tabular}{lll}
    \toprule
    Software & Recommended graphics device \\
    \midrule
    Illustrator & svg \\
    latex & ps \\
    MS Office & png (600 dpi) \\
    Open Office & png (600 dpi) \\
    pdflatex & pdf, png (600 dpi) \\
    web & png (72 dpi) \\
    \bottomrule 
  \end{tabular}
  \end{center}
  \caption{Recommended graphic output for different purposes.}
  \label{tbl:graphic-recommendation}
\end{table}

## Multiple plots on the same page {#sec:grid-layout}

If you want to arrange multiple plots on a single page, you'll need to learn a little bit of grid, the underlying graphics system used by ggplot. The key concept you'll need to learn about is a viewport: a rectangular subregion of the display. The default viewport takes up the entire plotting region, and by customising the viewport you can arrange a set of plots in just about any way you can imagine. \index{Layout} \index{Publication!multiple plots on the same page}

To begin, let's create three plots that we can experiment with. When arranging multiple plots on a page, it will usually be easiest to create them, assign them to variables and then plot them. This makes it easier to experiment with plot placement independent of content. The plots created by the code below are shown in Figure \ref{fig:layout}.

`r columns(3)`
```{r layout}
a <- ggplot(economics, aes(date, unemploy)) + geom_line()
b <- ggplot(economics, aes(uempmed, unemploy)) + geom_point() + geom_smooth(se = FALSE) 
c <- ggplot(economics, aes(uempmed, unemploy)) + geom_point()
a
b
c
```

### Subplots

One common layout is to have a small subplot drawn on top of the main plot.  To achieve this effect, we first plot the main plot, and then draw the subplot in a smaller viewport.  Viewports are created with (surprise!) the `viewport()` function, with parameters `x`, `y`, `width` and `height` to control the size and position of the viewport. By default, the measurements are given in 'npc' units, which range from 0 to 1. The location (0, 0) is the bottom left, (1, 1) the top right and (0.5, 0.5) the centre of viewport.  If these relative units don't work for your needs, you can also use absolute units, like `unit(2, "cm")` or `unit(1, "inch")`. \index{Sub-figures} \index{Subplots}

```{r viewports}
library(grid)
# A viewport that takes up the entire plot device
vp1 <- viewport(width = 1, height = 1, x = 0.5, y = 0.5)
vp1 <- viewport()

# A viewport that takes up half the width and half the height, 
# located in the middle of the plot.
vp2 <- viewport(width = 0.5, height = 0.5, x = 0.5, y = 0.5)
vp2 <- viewport(width = 0.5, height = 0.5)

# A viewport that is 2cm x 3cm located in the center
vp3 <- viewport(width = unit(2, "cm"), height = unit(3, "cm"))
```

By default, the x and y parameters control the location of the centre of the viewport. When positioning the plot in other locations, you may need to use the `just` parameter to control which corner of the plot you are positioning.  The following code gives some examples.

```{r viewport2}
# A viewport in the top right
vp4 <- viewport(x = 1, y = 1, just = c("right", "top"))
# Bottom left
vp5 <- viewport(x = 0, y = 0, just = c("right", "bottom"))
```

To draw the plot in our new viewport, we use the `vp` argument of the `print()` method.  This method is normally called automatically whenever you evaluate something on the command line, but because we want to customise the viewport, we need to call it ourselves. The result of this is shown in Figure \ref{fig:subplot-1}.

```{r polish, results='hide'}
pdf("figures/polishing-subplot-1.pdf", width = 4, height = 4)
subvp <- viewport(width = 0.4, height = 0.4, x = 0.75, y = 0.35)
b
print(c, vp = subvp)
dev.off()
```

This gives us what we want, but we need to make a few tweaks to the appearance: the text should be smaller, we want to remove the axis labels and shrink the plot margins. The result is shown in Figure \ref{fig:subplot-2}.

```{r polish2, results='hide'}
csmall <- c + 
  theme_gray(9) + 
  labs(x = NULL, y = NULL) + 
  theme(plot.margin = unit(c(1/4, 0, 0, 0), "lines"))

pdf("figures/polishing-subplot-2.pdf", width = 4, height = 4)
b
print(csmall, vp = subvp)
dev.off()
```

\begin{figure}[htbp]
  \centering
  \subfigure[Figure with subplot.]{
    \includegraphics[width=0.5\textwidth]{figures/polishing-subplot-1}
    \label{fig:subplot-1}
  }%
  \subfigure[Subplot tweaked for better display.]{
    \includegraphics[width=0.5\textwidth]{figures/polishing-subplot-2}
    \label{fig:subplot-2}
  }
  \caption{Two examples of a figure with subplot. It will usually be necessary to tweak the theme settings of the subplot for optimum display.}
  \label{fig:subplot}
\end{figure}

Note we need to use `pdf()` (or `png()` etc.) to save the plots to disk because `ggsave()` only saves a single plot.

### Rectangular grids

A more complicated scenario is when you want to arrange a number of plots in a rectangular grid. Of course you could create a series of viewports and use what you've learned above, but doing all the calculations by hand is cumbersome. A better approach is to use `grid.layout()`, which sets up a regular grid of viewports with arbitrary heights and widths. You still need to create each viewport, but instead of explicitly specifying the position and size, you can specify the row and column of the layout.

The following example shows how this work.  We first create the layout, here a 2 by 2 grid, then assign it to a viewport and push that viewport on to the plotting device.  Now we are ready to draw each plot into its own position on the grid.  We create a small function to save some typing, and then draw each plot in the desired place on the grid.  You can supply a vector of rows or columns to span a plot over multiple cells.  The results are shown in Figure \ref{fig:layout-2}.

`r columns(1)`
```{r layout-2}
grid.newpage()
pushViewport(viewport(layout = grid.layout(2, 2)))

vplayout <- function(x, y) 
  viewport(layout.pos.row = x, layout.pos.col = y)
print(a, vp = vplayout(1, 1:2))
print(b, vp = vplayout(2, 1))
print(c, vp = vplayout(2, 2))
```

By default `grid.layout()` makes each cell the same size, but you can use the `widths` and `heights` arguments to make them different sizes.  See the documentation for `grid.layout()` for more examples.

