---
title: qplot
output: bookdown::html_chapter
bibliography: references.bib
---

# Getting started with ggplot2 {#cha:qplot}

## Introduction

The goal of this chapter is to teach you how to produce useful graphics with ggplot2 as quickly as possible. You'll learn the basics of `ggplot()` and how to use it to make a wide variety of plots. `ggplot()` allows you to make complex plots with just a few lines of code because it's based on rich underlying theory, the grammar of graphics. You'll learn how to use the full expressive power of the grammar in later chapters. Here, we'll focus on some basic "recipes" so you can start using ggplot2 with a minimum of fuss.

In this chapter you'll learn:

* About the mpg mpg dataset included with ggplot2, [mpg](#mpg).

* The three key components of every plot: data, aesthetics and geoms;
  [key components](#sec:basic-use).
  
* How to add additional variables to a plot with aesthetics, 
  [aesthetics](#aesthetics).
  
* How to add addtional categorical variables to a plot by using
  small multiples created by facetting, [faceting](#sec:qplot-faceting).

* A variety of different geoms that you can use to create different
  types of plots, [geoms](#sec:plot-geoms).

## Fuel economy data {#sec:fuel-economy-data}

In this chapter we'll focus mostly on one data set that's bundled with ggplot2: `mpg`. It includes information about the fuel economy of popular car models in 1999 and 2008, collected by the US Environment Protection Agency, <http://fueleconomy.gov>.

You can access the data by loading ggplot2:

```{r}
library(ggplot2)
dim(mpg)
head(mpg)
```

The variables are mostly self-explanatory:

* `cty` and `hwy` record miles per gallon (mpg) for city and highway driving.
* `displ` is the engine displacement in litres
* `drv` is the drive train: front wheel (f), rear wheel (r) or four wheel (4)

It contains 38 models, selected because they had an update every year between 1999 and 2008. Most of the models should look familiar. \index{Data!mpg@\texttt{mpg}}

This dataset suggests many interesting questions. How are engine size and fuel economy related?  Do certain manufacturers care more about economy than others? Has fuel economy improved in the last ten years? We will try to answer some of these questions and in the process learn how to create some basic plots with ggplot2.

### Exercises

1.  What are five basic R tools that you could use to get more information 
    about this dataset?
    
1.  Apart from the US, most countries use fuel consumption (amount of fuel 
    needed to travel a fixed distance) rather than fuel economy (distance that
    fixed amount of fuel gets you). How could you convert `cty` and `hwy` into
    the European standard l/100km? 

## Key components {#sec:basic-use}

Every ggplot2 plot has three key components: some __data__, a set of __aesthetic mappings__ between variables in the data and visual properties, at least one __geom__, or geometric object.

Here's a simple example: \index{Scatterplot}

```{r qscatter}
ggplot(mpg, aes(x = displ, y = hwy)) + 
  geom_point()
```

This produces a scatterplot with:

* Data: `mpg`.
* Aesthetic mapping: engine displacement mapped to x position, 
  fuel economy to y position.
* Geom: points.

Pay attention to the structure of this function call: data and aesthetic mappings are supplied in `ggplot()`, then geoms are added on with `+`. This is an important pattern and as you learn more about ggplot2 you'll construct increasingly sophisticated plots by adding on more types of components.

Almost every plot maps a variable to `x` and `y`, so naming these aesthetics is going to get tedious fast. For that reason, the first two unnamed arguments to `aes()` will be mapped to `x` and `y`. That means that following code is identical to the example above:

```{r, eval = FALSE}
ggplot(mpg, aes(displ, hwy) + 
  geom_point()
```

I'll stick to that style throughout the book, so remember that the first two arguments are mapped to position.

The plot shows a strong correlation: as the engine size gets bigger, the fuel economy gets worse. There are also some interesting outliers: some cars with large engines get higher fuel economy than average.  What sort of cars do you think they are?

### Exercises

1.  How would you describe the relationship between `cty` and `hwy`?
  
1.  What does `ggplot(mpg, aes(model, manufacturer)) + geom_point()` show?
    Is it useful? How could you modify the data to more it more informative?
  
1.  Describe the data, aesthetic mappings and layers used for each of the 
    following plots. You'll need to guess a little because you haven't learned
    what all of the functions do; use your common sense! See if you can
    predict what the plot will look like before running the code.
    
    1. `ggplot(mpg, aes(cty, hwy)) + geom_point()`
    1. `ggplot(diamonds, aes(carat, price)) + geom_point()`
    1. `ggplot(economics, aes(date, unemploy)) + geom_line()`
    1. `ggplot(mpg, aes(cty)) + geom_histogram()`

## Colour, size, shape and other aesthetic attributes {#aesthetics}

To add additional variables to a plot, we can use other aesthetics like colour, shape, and size. These work just like `x` and `y`, you just add them into the `aes()` call:

* `aes(displ, hwy, colour = class)`
* `aes(displ, hwy, shape = drv)`
* `aes(displ, hwy, size = cyl)`

(Note that while I use British spelling throughout this book, ggplot2 also accepts American spellings.)

ggplot2 takes care of the details of converting a categorical variable in your data (e.g., 'f', 'r', '4') into something that the graphics device knows how to display (e.g., 'red', 'yellow', 'green'). It also adds a legend gives the inverse mapping, showing how the displayed attributes can be converted back to the data values. \index{Aesthetics}

To learn more about those outlying variables in the previous scatterplot, we could map the class variable to colour:

```{r qplot-aesthetics}
ggplot(mpg, aes(displ, cty, colour = class)) + 
  geom_point()
```

This gives each point a unique colour corresponding to its class. The legend allows us to map back from colour to data value, shows us that the group of cars with unusually high fuel economy for their engine size are two-seaters: cars with big engines, but lightweight bodies. 

For every aesthetic attribute, there is a function, called a _scale_, which maps data values to valid values for that aesthetic. It is this scale that controls the appearance of the points and associated legend. For example, in the above plot, the colour scale maps 2seater to pink and midsize to green. You can easily override ggplot2's default choices, as you'll learn in [the scales chapter](#cha:scales). 

If you want to set an aesthetic to a fixed value, without scaling it, do so in the individual geom. Compare the following two plots:

```{r}
ggplot(mpg, aes(displ, hwy, colour = "blue")) + 
  geom_point()
ggplot(mpg, aes(displ, hwy)) + 
  geom_point(colour = "blue")
```

In the first plot, the value "blue" is scaled to a pinkish colour, and a legend is added. In the second plot, the points are given the R colour blue. This is an important technique and is explained in more detail in [setting vs. mapping](#sub:setting-mapping).

Different types of aesthetic attributes work better with different types of variables. For example, colour and shape work well with categorical variables, while size works better with continuous variables. The amount of data also makes a difference: if there is a lot of data, like in the plots above, it can be hard to distinguish the different groups. An alternative solution is to use faceting, as described next.

When using aesthetics in a plot, less is usually more. It's difficult to see the simultaneously relationship between colour and size and size, so exercise restraint when using aesthetics. Instead trying to make one very complex plot that shows everything, see if you can create a series of simple plots that tell a story.

### Exercises

1.  Experiemnt with the colour, shape and size aesthetics. What happens
    if you use more than one aesthetic in a plot?

1.  What happens if you map a continuous variable to shape? Why? What happens
    if you map trans to shape? Why?
    
1.  How does the drive train related to fuel economy? How is it 
    related to engine size and class?

## Faceting {#sec:qplot-faceting}

Another way of displaying additional categorical variables on a plot is facetting. Faceting creates tables of graphics by splitting the data into subsets and displaying the same graph for each subset in an arrangement that facilitates comparison. You'll learn more about faceting in [Faceting](#sec:faceting), but it's such a useful technique that it's worth learning right away.  \index{Faceting}. You'll learn more about the relative advantages and disadvantages of using faceting instead of aesthetics in [grouping vs. faceting](#sub:group-vs-facet).

There are two types of facetting: grid and wrapped. Wrapped is the most useful, so we'll discuss it here, and you can learn about grid facetting later. To facet a plot you simply add a facetting specification:

```{r facet}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  facet_wrap(~class)
```

`facet_wrap()` takes the name of a variable preceded by `~`. (You'll learn why it uses this special syntax later). \indexf{facet_wrap}

### Exericses

1.  What happens if you try to facet by a continuous variable like 
    `hwy`? What about `cyl`?  What's the key difference?
  
1.  Use facetting to explore the 3-way relationship between fuel economy,
    engine size and number of cylinders. How does facetting by number of
    cylinders change your assessement of the relationship between
    engine size and fuel economy?

1.  Read the documentation for `facet_wrap()`. What arguments can you use
    to control how many rows and columns in the output?

## Plot geoms {#sec:plot-geoms}

You might guess that by substituting `geom_point()` for a different geom function, you'd get a different plot. That's a great guess!

Some geoms have an associated statistical transformation, for example, a histogram is a binning statistic plus a bar geom. Those different components are described in the next chapter. Here we'll introduce the most common and useful geoms, organised by the dimensionality of data that they work with. The following geoms enable you to investigate two-dimensional relationships:

* `geom_smooth()` fits a smoother to the data and displays the smooth and its 
  standard error, see [adding a smoother to a plot](#sub:smooth).

* `geom_boxplot()` produces a box-and-whisker plot to summarise the distribution 
  of a set of points, see [boxplots and jittered points](#sub:boxplot).

* `geom_path()` and `geom_line()` draw lines between the data points.
  A line plot is constrained to produce lines that travel from left to right, 
  while paths can go in any direction, see 
  [time series with line and path plots](#sub:line). Lines are typically used 
  to explore relationships between time and another variable, and paths to join 
  observations connected in some other way.  

For 1d distributions, your choice of geoms is guided by the variable type:

* For continuous variables, `geom_histogram()` draws a histogram, and
  `geom_freqpoly()` a frequency polygon, 
  [histogram and frequency polygons](#sub:distribution).

* For discrete variables, use `geom_bar()` to make a bar chart, 
  [bar charts](#sub:bar).

## Adding a smoother to a plot {#sub:smooth}

If you have a scatterplot with many data points, it can be hard to see exactly what trend is shown by the data. In this case you may want to add a smoothed line to the plot. This is easily done using `geom_smooth()`: \index{Smoothing} \indexf{geom_smooth}

```{r qplot-smooth}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  geom_smooth()
```

This overlays the scatterplot with a smooth curve, including an assessment of uncertainty in the form point-wise confidence interval shown in grey. If you're not interested in the confidence interval, turn it off with `geom_smooth(se = FALSE)`.

Another important argument to `geom_smooth()` is method, which allows you to choose the type of model used for smoothing:

*   `method = "loess"`, the default for small n, uses a smooth local 
    regression.  More details about the algorithm used can be found in `?loess`.  
    The wiggliness of the line is controlled by the `span` parameter, which 
    ranges from 0 (exceedingly wiggly) to 1 (not so wiggly). \index{Model!loess}
      
    ```{r smooth-loess, fig.show = "hold"}
    ggplot(mpg, aes(displ, hwy)) + 
      geom_point() + 
      geom_smooth(span = 0.2)

    ggplot(mpg, aes(displ, hwy)) + 
      geom_point() + 
      geom_smooth(span = 1)
    ```
      
    Loess does not work well for large datasets (it's $O(n^2)$ in memory), so
    an alternative smoothing algorithm is used when $n$ is greater than 1,000. 

*   You could load the __mgcv__ package and use `method = "gam"` and 
    `formula = y ~ s(x)` to fit a generalised additive model. For large data, 
    use the formula `y ~ s(x, bs = "cs")`. This is what ggplot2 uses when 
    there are more than 1,000 points. \index{Package!mgcv}
    \index{Model!generalised additive}

    ```{r smooth-gam}
    library(mgcv)
    ggplot(mpg, aes(displ, hwy)) + 
      geom_point() + 
      geom_smooth(method = "gam", formula = y ~ s(x))
    ```

*   `method = "lm"` fits a linear model, giving the line of best fit.     
    \index{Model!linear}

    ```{r smooth-lm, fig.show = "hold"}
    ggplot(mpg, aes(displ, hwy)) + 
      geom_point() + 
      geom_smooth(method = "lm")
    ```

  * `method = "rlm"` works like `lm()`, but uses a robust fitting algorithm so 
    that outliers don't affect the fit as much.  It's part of the __MASS__ 
    package, so remember to load that first. \index{Model!robust} 
    \index{Package!MASS}

### Exercises

## Boxplots and jittered points {#sub:boxplot}

When a set of data includes a categorical variable and one or more continuous variables, you will probably be interested to know how the values of the continuous variables vary with the levels of the categorical variable. There are three useful techniques that help see the distribution:

* Jittering, `geom_jitter()`, adds a little random noise to the data.
  \index{Jittering}  \indexf{geom_jitter}
  
* Boxplots, `geom_boxplot()`, summarise the shape of the distribution
  with a handful of summary statistics. \index{Boxplot}  \indexf{geom_boxplot}
  
* Violin plots, `geom_violin()`, show a compact representation of the
  "density" of the distribution, highlighting the areas where more points
  are found.

Say we're interested in seeing how fuel economy varies within car class. We might start with a scatterplot like this:

```{r}
ggplot(mpg, aes(class, hwy)) + 
  geom_point()
```

But because there are few unique values of both class and hwy, there is a lot of overplotting. Many points are plotted in the same location, and it's difficult to see the distribution.

```{r jitter-boxplot, fig.show = "hold"}
ggplot(mpg, aes(class, hwy)) + 
  geom_jitter()

ggplot(mpg, aes(class, hwy)) + 
  geom_boxplot()

ggplot(mpg, aes(class, hwy)) + 
  geom_violin()
```

Each method has its strengths and weaknesses. Boxplots summarise the bulk of the distribution with only five numbers, while jittered plots show every point but can suffer from overplotting. Violin plots are the richest, but relying on the calculation of a density estimate, which can be hard to interpret.

For jittered points, `geom_jitter()` offers the same control over aesthetics as `geom_point()`: `size`, `colour`, `shape` and `alpha`. For `geom_boxplot()` and `geom_violin()`, you can control the outline `colour`, the internal `fill` colour and the `size` of the lines.

### Exercises

1.  What's the problem with the plot created by 
    `ggplot(mpg, aes(cty, hwy)) + geom_point()`? Which of the three geoms 
    described above is most effective a remedying?

1.   One challenge with the plots shown above is that the ordering of
    `class` is alphabetic, and not terrible useful. How would you change the 
    levels of the factor to be most informative? 
   
    Rather than reordering the factor by hand, you can do it automatically
    based on the data:
    `ggplot(mpg, aes(reorder(class, hwy), hwy)) + geom_boxplot()`.
    What does `reorder()` do? Read the documentation.

## Histograms and frequency polygons {#sub:distribution}

Histograms and freqency polygons show the distribution of a single variable. They provide more information about the distribution of a single group than boxplots do, but also take up more space so it's harder to show many groups at once. \index{Histogram} \indexf{geom_histogram}

```{r dist, fig.show = "hold", fig.width = 4, fig.height = 3}
ggplot(mpg, aes(hwy)) + 
  geom_histogram()
ggplot(mpg, aes(hwy)) + 
  geom_freqpoly()
```

Both histograms and frequency polygons work in the same way: they divide the data up into bins and count the number of observations in each bin. The only difference is the display: histograms use bars and frequency polygons use lines. You can how the data is divided into bins with the `binwidth` argument (if you don't want evenly spaced bins you can use the `breaks` argument). It is __very important__ to experiment with the bin width. The default just splits your data into 30 bins, which is unlikely to be the best choice. You should always try many bin widths: You may find that gross features of the data show up well at a large bin width, while finer features require a very narrow width.

```{r}
ggplot(mpg, aes(hwy)) + 
  geom_freqpoly(binwidth = 1)
ggplot(mpg, aes(displ)) + 
  geom_freqpoly(binwidth = 0.5)
```

(An alternative to the frequency polygon is the density plot, `geom_density()`. I generally avoid density plots because they are harder to interpret since the underlying computations are more complex. They also make assumptions that are not true for all data, namely that the underlying distribution is continuous, unbounded, and smooth.)

To compare the distributions of different subgroups, you can map a categorical variable to either fill (for `geom_histogram()`) or colour (for `geom_freqpoly()`). It's better to compare distributions with using the frequency polygon because the perceptual task is easier: you can compare the heights against a common baseline.

```{r dist-fill, fig.show = "hold", fig.width = 4, fig.height = 3}
ggplot(mpg, aes(displ, colour = drv)) + 
  geom_freqpoly(binwidth = 0.5)
ggplot(mpg, aes(displ, fill = drv)) + 
  geom_histogram(binwidth = 0.5)
```

### Exercises

1.  Explore the distribution of the carat variable in the dimaonds 
    dataset. What binwidth reveals the most interesting patterns?
    
1.  Explore the distribution of the price variable in the diamonds
    data. How does the distribution vary by cut.

1.  You now know three ways to compare the distributions of subgroups:
    with `geom_violin()`, with `geom_freqpoly()` and the colour aesthetic,
    and with `geom_freqpoly()` and faceting. What are the strengths 
    and weaknesses of each approach?

## Bar charts {#sub:bar}

The discrete analogue of the histogram is the bar chart, `geom_bar()`. It's easy to use: \index{Barchart} \indexf{geom_bar}

```{r dist-bar}
ggplot(mpg, aes(manufacturer)) + 
  geom_bar()
```

Bar charts can be confusing because there are two rather different plots that are both commonly called bar charts. The above form expects you to have unsummarised data - each row contributes 1 to the overall count. The other form of bar chart is when you have presummarised data. For you might have three drugs along with their mean effect:

```{r}
drugs <- data.frame(
  drug = c("a", "b", "c"),
  effect = c(4.2, 9.7, 6.1)
)
```

To display this sort of data, you need to tell `geom_bar()` to not run the default statistical transformation which bins and counts the data:.

```{r}
ggplot(drugs, aes(drug, effect)) +
  geom_bar(stat = "identity")
```

But if you have this sort of data, you should seriously consider using `geom_point()` instead.

### Exercises

1.  Read the documentation for `geom_bar()`. What does the `weight` 
    aesthetic do? 

1.  Using the techniques already discusssed in this chapter, come up with
    three ways to visualise of 2 categorical variables simultaneously.
    Try them out to visualise the distribution of `model` and `manufacturer`, 
    `trans` and `class`, and `cyl` and `trans`.

## Time series with line and path plots {#sub:line}

Line and path plots are typically used for time series data. Line plots join the points from left to right, while path plots join them in the order that they appear in the dataset (in other words, a line plot is a path plot of the data sorted by x value). Line plots usually have time on the x-axis, showing how a single variable has changed over time. Path plots show how two variables have simultaneously changed over time, with time encoded in the way that observations are connected.

Because the year variable in the mpg dataset only has twov values, we'll show some timeseries plots using the `economics` dataset, which contains economic data on the US measured over the last 40 years. The figure below shows two plots of unemployment over time, both produced using `geom_line()`. The first shows an unemployment rate and the second shows the median number of weeks unemployed. We can already see some differences in these two variables, particularly in the last peak, where the unemployment percentage is lower than it was in the preceding peaks, but the length of unemployment is high. \index{Time series!bivariate} \indexf{geom_line} \indexf{geom_path}

```{r line-employment, fig.show = "hold", fig.width = 4, fig.height = 2.4}
ggplot(economics, aes(date, unemploy / pop)) +
  geom_line()
ggplot(economics, aes(date, uempmed)) +
  geom_line()
```

To examine this relationship in greater detail, we would like to draw both time series on the same plot. We could draw a scatterplot of unemployment rate vs. length of unemployment, but then we could no longer see the evolution over time. The solution is to join points adjacent in time with line segments, forming a _path_ plot.

Below we plot unemployment rate vs. length of unemployment and join the individual observations with a path. Because of the many line crossings, the direction in which time flows isn't easy to see in the first plot. In the second plot, we colour the points to make it easier to see the direction of time.

```{r path-employ, fig.show = "hold", fig.width = 4, fig.height = 2.4}
ggplot(economics, aes(unemploy / pop, uempmed)) + 
  geom_path() +
  geom_point()

year <- function(x) as.POSIXlt(x)$year + 1900
ggplot(economics, aes(unemploy / pop, uempmed)) + 
  geom_path(colour = "grey50") +
  geom_point(aes(colour = year(date))) 
```

We can see that percent unemployed and length of unemployment are highly correlated, but in recent years the length of unemployment has been increasing relative to the unemployment rate.

With longitudinal data, you often want to display multiple time series on each plot, each series representing one individual. To do this you need to map the `group` aesthetic to a variable encoding the group membership of each observation. This is explained in more depth in [grouping](#sub:grouping). \index{Longitudinal data|see{Time series}} \index{Data!longitudinal}
