---
title: qplot
output: bookdown::html_chapter
bibliography: references.bib
---

# Getting started with qplot {#cha:qplot}

## Introduction

In this chapter, you will the basics of the grammar underlying ggplot2. You'll learn to make a wide variety of plots with your first ggplot function, `ggplot()`. `ggplot()` makes it easy to produce complex plots, often requiring many lines of code using other plotting systems, in one line. `ggplot()` can do this because it's based on the grammar of graphics, which allows you to create a simple, yet expressive, description of the plot. In later chapters you'll learn to use all of the expressive power of the grammar, but here we'll start simple so you can work your way up.  You will also start to learn some of the ggplot terminology that will be used throughout the book.

In this chapter you'll learn:

* The basic use of `qplot()`---If you're already familiar with `plot()`, this 
  will be particularly easy, [link to section](#sec:basic-use).

* How to map variables to aesthetic attributes, like colour, size and shape, 
  [link to section](#sec:aesthetic-attributes).

* How to create many different types of plots by specifying different geoms, 
  and how to combine multiple types in a single plot, 
  [link to section](#sec:plot-geoms).

* The use of faceting, also known as trellising or conditioning, to break 
  apart subsets of your data, [link to section](#sec:qplot-faceting).

## Diamonds data {#sec:data-sets}

```{r}
library(ggplot2)
```

In this chapter we'll focus primarily on one data source. The `diamonds` dataset consists of prices and quality information about 54,000 diamonds, and is included in the __ggplot2__ package. The data contains the four C's of diamond quality, carat, cut, colour and clarity; and five physical measurements, depth, table, x, y and z, as described in Figure~\ref{fig:diamonds}. The first few rows of the data are shown below. \index{Data!diamonds@\texttt{diamonds}}

```{r diamonds, echo = FALSE}
knitr::kable(head(diamonds))
```

\begin{figure}[htbp]
  \centering
    \includegraphics[width=0.8\linewidth]{diagrams/diamond-dimensions}
  \caption{How the variables x, y, z, table and depth are measured.}
  \label{fig:diamond-dim}
\end{figure}

The dataset has not been well cleaned, so as well as demonstrating interesting relationships about diamonds, it also demonstrates some data quality problems. We'll also use another dataset, `dsmall`, which is a random sample of 100 diamonds. We'll use this data for plots that are more appropriate for smaller datasets. 

```{r dsmall}
set.seed(1410) # Make the sample reproducible
dsmall <- diamonds[sample(nrow(diamonds), 100), ]
```

## Basic use {#sec:basic-use}

data, aesthetics with aes, +, geom_point.

Here is a simple example of the use of `ggplot2()`.  It produces a scatterplot showing the relationship between the price and carats (weight) of a diamond.  \index{Scatterplot}

```{r qscatter, dev="png"}
ggplot(diamonds, aes(carat, price)) + 
  geom_point()
```

The plot shows a strong correlation with notable outliers and some interesting vertical striation.  The relationship looks exponential, though, so the first thing we'd like to do is to transform the variables. Because `qplot()` accepts functions of variables as arguments, we plot `log(price)` vs. `log(carat)`:

```{r qlog, dev="png"}
ggplot(diamonds, aes(log(carat), log(price))) + 
  geom_point()
```

The relationship now looks linear.  With this much overplotting, though, we need to be cautious about drawing firm conclusions.

Arguments can also be combinations of existing variables, so, if we are curious about the relationship between the volume of the diamond (approximated by $x \times y \times z$) and its weight, we could do the following:

```{r volume, dev="png"}
ggplot(diamonds, aes(carat, x * y * z)) + 
  geom_point()
```

We would expect the density (weight/volume) of diamonds to be constant, and so see a linear relationship between volume and weight. The majority of diamonds do seem to fall along a line, but there are some large outliers.

## Colour, size, shape and other aesthetic attributes {#sec:aesthetic-attributes}

To additional variables to a plot, we need to use aesthetics. ggplot2 takes care of the details of converting a categorical variable in your data (e.g., 'apples', 'bananas', 'pears') into something that `plot()` knows how to use (e.g., 'red', 'yellow', 'green').  `qplot()` can do this for you automatically, and it will automatically provide a legend that maps the displayed attributes to the data values.  This makes it easy to include additional data on the plot.  

Here we augment the plot of carat and price with information about diamond colour and cut:

```{r qplot-aesthetics, fig.show = "hold"}
ggplot(dsmall, aes(carat, price)) + 
  geom_point(aes(colour = color))
ggplot(dsmall, aes(carat, price)) + 
  geom_point(aes(colour = cut))
```

Colour, size and shape are all examples of aesthetic attributes, visual properties that affect the way observations are displayed. For every aesthetic attribute, there is a function, called a _scale_, which maps data values to valid values for that aesthetic. It is this scale that controls the appearance of the points and associated legend. For example, in the above plots, the colour scale maps J to purple and F to green. (Note that while I use British spelling throughout this book, the software also accepts American spellings.)\index{Aesthetics}

You can also manually set the aesthetics by placing them outside of `aes()`.  This is not the same as mapping and is explained in more detail in [setting vs. mapping](#sub:setting-mapping).  For large datasets, like the diamonds data, semi-transparent points are often useful to alleviate some of the overplotting.  To make a semi-transparent colour you can use the alpha aesthetic, which takes a value between 0 (completely transparent) and 1 (complete opaque).  It's often useful to specify the transparency as a fraction, e.g., `1/10` or `1/20`, as the denominator specifies the number of points that must overplot to get a completely opaque colour. \index{Aesthetics!setting} \indexf{I}

```{r qplot-set, fig.show = "hold", dev = "png", out.width = "0.32\\linewidth", fig.weight = 4, fig.height = 4}
ggplot(diamonds, aes(carat, price)) + 
  geom_point(alpha = 1 / 10)
ggplot(diamonds, aes(carat, price)) + 
  geom_point(alpha = 1 / 100)
ggplot(diamonds, aes(carat, price)) + 
  geom_point(alpha = 1 / 200)
```

Different types of aesthetic attributes work better with different types of variables. For example, colour and shape work well with categorical variables, while size works better with continuous variables. The amount of data also makes a difference: if there is a lot of data, like in the plots above, it can be hard to distinguish the different groups. An alternative solution is to use faceting (described in [link to section](#sec:qplot-faceting)).

### Exercises

1.  What happens if you set alpha smaller than 1/1000? Why?

## Plot geoms {#sec:plot-geoms}

`?geom_smooth`.

`ggplot()` is not limited to scatterplots, but can produce almost any kind of plot by varying the `geom`. Geom, short for geometric object, describes the type of object that is used to display the data. Some geoms have an associated statistical transformation, for example, a histogram is a binning statistic plus a bar geom. These different components are described in the next chapter. Here we'll introduce the most common and useful geoms, organised by the dimensionality of data that they work with. The following geoms enable you to investigate two-dimensional relationships:

* `geom_smooth()` fits a smoother to the data and displays the smooth and its 
  standard error, see [adding a smoother to a plot](#sub:smooth).

* `geom_boxplot()` produces a box-and-whisker plot to summarise the distribution 
  of a set of points, see [boxplots and jittered points](#sub:boxplot).

* `geom_path()` and `geom_line()` draw lines between the data points.
  A line plot is constrained to produce lines that travel from left to right, 
  while paths can go in any direction, see 
  [time series with line and path plots](#sub:line). Lines are typically used 
  to explore relationships between time and another variable, and paths to join 
  observations connected in some other way.  

For 1d distributions, your choice of geoms is guided by the variable type:

* For continuous variables, `geom_histogram()` draws a histogram, and
  `geom_freqpoly()` a frequency polygon, 
  [histogram and frequency polygons](#sub:distribution).

* For discrete variables, use `geom_bar()` to make a bar chart, see 
  [bar charts](#sub:bar).

### Adding a smoother to a plot {#sub:smooth}

If you have a scatterplot with many data points, it can be hard to see exactly what trend is shown by the data. In this case you may want to add a smoothed line to the plot. This is easily done using the `smooth` geom as shown below. The geoms will be overlaid in the order in which they appear.  \index{Smoothing} \indexf{geom_smooth}

```{r qplot-smooth, fig.show = "hold", dev = "png"}
ggplot(diamonds, aes(carat, price)) + 
  geom_point() + 
  geom_smooth()
```

Despite overplotting, our impression of an exponential relationship between price and carat was correct. There are few diamonds bigger than three carats, and our uncertainty in the form of the relationship increases as illustrated by the point-wise confidence interval shown in grey. If you want to turn the confidence interval off, use `geom_smooth(se = FALSE)`.

There are many different smoothers you can choose between by using the `method` argument:

*   `method = "loess"`, the default for small n, uses a smooth local 
    regression.  More details about the algorithm used can be found in `?loess`.  
    The wiggliness of the line is controlled by the `span` parameter, which 
    ranges from 0 (exceedingly wiggly) to 1 (not so wiggly). \index{Model!loess}
      
    ```{r smooth-loess, fig.show = "hold"}
    ggplot(dsmall, aes(carat, price)) + 
      geom_point() + 
      geom_smooth(span = 0.2)
    
    ggplot(dsmall, aes(carat, price)) + 
      geom_point() + 
      geom_smooth(span = 1)
    ```
      
    Loess does not work well for large datasets (it's $O(n^2)$ in memory), so
    an alternative smoothing algorithm is used when $n$ is greater than 1,000. 

*   You could load the __mgcv__ library and use `method = "gam"` and 
    `formula = y ~ s(x)` to fit a generalised additive model. For large data, 
    use the formula `y ~ s(x, bs = "cs")`. This is used by default when there 
    are more than 1,000 points. 
    \index{Package!mgcv} \index{Model!generalised additive}

    ```{r smooth-gam}
    library(mgcv)
    ggplot(dsmall, aes(carat, price)) + 
      geom_point() + 
      geom_smooth(method = "gam", formula = y ~ s(x))
    ```

*   `method = "lm"` fits a linear model.  The default will fit a straight line 
    to your data, or you can specify `formula = y ~ poly(x, 2)` to specify a 
    2nd-degree polynomial, or better, load the __splines__ package and use a 
    natural spline: `formula = y ~ ns(x, 2)`. The second parameter is the 
    degrees of freedom: a higher number will create a wigglier curve. You are 
    free to specify any formula involving $x$ and $y$.  \index{Model!linear}

    ```{r smooth-lm, fig.show = "hold"}
    library(splines)
    
    # A linear model
    ggplot(dsmall, aes(carat, price)) + 
      geom_point() + 
      geom_smooth(method = "lm")
    
    # A natural spline with 5 degrees of freedom
    ggplot(dsmall, aes(carat, price)) + 
      geom_point() + 
      geom_smooth(method = "lm", formula = y ~ ns(x, 5))
    ```

  * `method = "rlm"` works like `lm()`, but uses a robust fitting algorithm so 
    that outliers don't affect the fit as much.  It's part of the __MASS__ 
    package, so remember to load that first. \index{Model!robust} 
    \index{Package!MASS}

### Boxplots and jittered points {#sub:boxplot}

When a set of data includes a categorical variable and one or more continuous variables, you will probably be interested to know how the values of the continuous variables vary with the levels of the categorical variable.  Box-plots and jittered points offer two ways to do this.  The following code explores how the distribution of price per carat varies with the colour of the diamond using jittering (`geom = "jitter"`, left) and box-and-whisker plots (`geom = "boxplot"`, right). \index{Boxplot} \index{Jittering} \indexf{geom_boxplot}

```{r jitter-boxplot, fig.show = "hold", dev = "png"}
# As the colour improves (from left to right) the spread of values decreases, 
# but there is little change in the centre of the distribution.
ggplot(diamonds, aes(color, price / carat)) + 
  geom_jitter()

ggplot(diamonds, aes(color, price / carat)) + 
  geom_boxplot()

ggplot(diamonds, aes(color, price / carat)) + 
  geom_violin()
```

Each method has its strengths and weaknesses. Boxplots summarise the bulk of the distribution with only five numbers, while jittered plots show every point but can suffer from overplotting. In the example here, both plots show the dependency of the spread of price per carat on diamond colour, but the boxplots are more informative, indicating that there is very little change in the median and adjacent quartiles.

For jittered points, `geom_jitter()` offers the same control over aesthetics as `geom_point()`: `size`, `colour`, `shape` and `alpha`. For `geom_boxplot()` and `geom_violin()`, you can control the outline `colour`, the internal `fill` colour and the `size` of the lines.

Another way to look at conditional distributions is to use faceting (described in [faceting](#sec:qplot-faceting)) to plot a separate histogram or density plot for each value of the categorical variable.

### Histograms and frequency polygons {#sub:distribution}

Histogram and freqency polygons show the distribution of a single variable.  They provide more information about the distribution of a single group than boxplots do, but it is harder to compare many groups (although we will look at one way to do so). \index{Histogram} \index{Density!plot} \indexf{geom_histogram} \indexf{geom_density}

```{r dist, fig.show = "hold", fig.width = 4, fig.height = 3}
ggplot(diamonds, aes(carat)) + 
  geom_histogram()
ggplot(diamonds, aes(carat)) + 
  geom_freqpoly()
```

Both both geoms, you can can controls the amount of smoothing by setting the bin width. (Break points can also be specified explicitly, using the `breaks` argument.) It is __very important__ to experiment with the bin width  You should always try many bin widths: You may find that gross features of the data show up well at a large bin width, while finer features require a very narrow width.

Below we experiment with three values of `binwidth`: 1.0, 0.1 and 0.01. It is only in the plot with the smallest bin width (right) that we see the striations we noted in an earlier scatterplot, most diamonds have a 'nice' numbers of carats:

```{r hist-binwidth, fig.show = "hold", fig.width = 4, fig.height = 3, out.width = "0.32\\linewidth"}
ggplot(diamonds, aes(carat)) + 
  geom_histogram(binwidth = 1) + 
  xlim(0, 3)

ggplot(diamonds, aes(carat)) + 
  geom_histogram(binwidth = 0.1) + 
  xlim(0, 3)

ggplot(diamonds, aes(carat)) + 
  geom_histogram(binwidth = 0.01) + 
  xlim(0, 3)
```

To compare the distributions of different subgroups, just add an aesthetic mapping, as in the following code:

```{r dist-fill, fig.show = "hold", fig.width = 4, fig.height = 3}
ggplot(diamonds, aes(price / carat)) + 
  geom_histogram(binwidth = 500)

ggplot(diamonds, aes(price / carat)) + 
  geom_histogram(aes(fill = color), binwidth = 500)
ggplot(diamonds, aes(price / carat)) + 
  geom_freqpoly(aes(color = color), binwidth = 500)
```

Mapping a categorical variable to an aesthetic will automatically split up the geom by that variable, so these commands instruct ggplot2 to draw a histogram and frequency polygon for each colour of diamond. It's easier to compare distributions when using the frequency polygon because the perceptual task is easier: you can compare the heights against a common baseline.

### Bar charts {#sub:bar}

The discrete analogue of histogram is the bar chart, `geom_bar()`. The bar geom counts the number of instances of each class so that you don't need to tabulate your values beforehand, as with `barplot()` in base R. If the data has already been tabulated or if you'd like to tabulate class members in some other way, such as by summing up a continuous variable, you can use the `weight` aesthetic. The first plot is a simple bar chart of diamond colour, and the second is a bar chart of diamond colour weighted by carat. \index{Barchart} \indexf{geom_bar}

```{r dist-bar, fig.show = "hold", fig.width = 4, fig.height = 2.4}
# Raw counts
ggplot(diamonds, aes(color)) + 
  geom_bar()

# Weighted by carat
ggplot(diamonds, aes(color)) + 
  geom_bar(aes(weight = carat)) + 
  scale_y_continuous("carat")
```

(Mention the other type of bar plot where the data is already summarised)

### Time series with line and path plots {#sub:line}

Line and path plots are typically used for time series data. Line plots join the points from left to right, while path plots join them in the order that they appear in the dataset (a line plot is just a path plot of the data sorted by x value).  Line plots usually have time on the x-axis, showing how a single variable has changed over time.  Path plots show how two variables have simultaneously changed over time, with time encoded in the way that the points are joined together.

Because there is no time variable in the diamonds data, we use the `economics` dataset, which contains economic data on the US measured over the last 40 years. The figure below shows two plots of unemployment over time, both produced using `geom = "line"`. The first shows an unemployment rate and the second shows the median number of weeks unemployed. We can already see some differences in these two variables, particularly in the last peak, where the unemployment percentage is lower than it was in the preceding peaks, but the length of unemployment is high. \index{Time series!bivariate} \indexf{geom_line} \indexf{geom_path}

```{r line-employment, fig.show = "hold", fig.width = 4, fig.height = 2.4}
ggplot(economics, aes(date, unemploy / pop)) +
  geom_line()
ggplot(economics, aes(date, uempmed)) +
  geom_line()
```

To examine this relationship in greater detail, we would like to draw both time series on the same plot. We could draw a scatterplot of unemployment rate vs. length of unemployment, but then we could no longer see the evolution over time. The solution is to join points adjacent in time with line segments, forming a _path_ plot.

Below we plot unemployment rate vs. length of unemployment and join the individual observations with a path. Because of the many line crossings, the direction in which time flows isn't easy to see in the first plot. In the second plot, we apply the `colour` aesthetic to the line to make it easier to see the direction of time.

```{r path-employ, fig.show = "hold", fig.width = 4, fig.height = 2.4}
ggplot(economics, aes(unemploy / pop, uempmed)) + 
  geom_path() +
  geom_point()

year <- function(x) as.POSIXlt(x)$year + 1900
ggplot(economics, aes(unemploy / pop, uempmed)) + 
  geom_path() +
  geom_point(aes(colour = year(date))) 
```

We can see that percent unemployed and length of unemployment are highly correlated, although in recent years the length of unemployment has been increasing relative to the unemployment rate.

With longitudinal data, you often want to display multiple time series on each plot, each series representing one individual. To do this you need to map the `group` aesthetic to a variable encoding the group membership of each observation. This is explained in more depth in [grouping](#sub:grouping). \index{Longitudinal data|see{Time series}} \index{Data!longitudinal}

## Faceting {#sec:qplot-faceting}

We have already discussed using aesthetics (colour and shape) to compare subgroups, drawing all groups on the same plot. Faceting takes an alternative approach: It creates tables of graphics by splitting the data into subsets and displaying the same graph for each subset in an arrangement that facilitates comparison. [Faceting](#sec:faceting) discusses faceting in detail, including a discussion of the advantages and disadvantages of using faceting instead of aesthetics in [grouping vs. faceting](#sub:group-vs-facet). \index{Faceting}

To add facetting you either use `facet_grid(row_var ~ col_var)` or `facet_wrap(~ var)`. To facet on only one of columns or rows, use `.` as a place holder.  For example, `row_var ~ .` will create a single column with multiple rows. \indexf{facet_grid}

The following code illustrates this technique with two plots, sets of histograms showing the distribution of carat conditional on colour. 

```{r facet-hist}
ggplot(diamonds, aes(price)) + 
  geom_histogram(binwidth = 500) + 
  facet_wrap(~color)
```

## mpg challenges

```{r}
knitr::kable(head(mpg))
```

### Exercises

1.  Facetting vs. aesthetics.

## `qplot()` shortcut


