```{r setup, include = FALSE}
source("common.R")
columns(1, 2 / 3)
```

# Position scales and axes {#scale-position}

Every plot has two position scales corresponding to the x and y aesthetics. Typically the user specifies the variables mapped to x and y explicitly, but sometimes an aesthetic is mapped to a computed variable, as happens with `geom_histogram()`, and does not need to be explicitly specified. For example, the following plot specifications are equivalent:

```{r, fig.show = "hide", message = FALSE}
ggplot(mpg, aes(x = displ)) + geom_histogram()
ggplot(mpg, aes(x = displ, y = after_stat(count))) + geom_histogram()
```

Although the first example does not state the y-aesthetic mapping explicitly, it still exists and is associated with (in this case) a continuous position scale.

\index{Scales!position} \index{Positioning!scales} 

<!-- HW: needs a bit more of an overview here. What am I going to learn from this chapter? How is it organised? -->

## Numeric position scales {#numeric-position-scales}

The most common continuous position scales are the default `scale_x_continuous()` and `scale_y_continuous()` functions. In the simplest case they map linearly from the data value to a location on the plot. There are several other position scales for continuous variables---`scale_x_log10()`, `scale_x_reverse()`, etc---most of which are convenience functions used to provide easy access to common transformations:

`r columns(3)`
```{r}
base <- ggplot(mpg, aes(displ, hwy)) + geom_point()

base
base + scale_x_reverse()
base + scale_y_reverse()
```

Scale transformations are discussed later in this chapter, in Section \@ref(scale-transformation).
\indexf{scale\_x\_continuous} 

<!-- HW: maybe a bit too much about position transformations here? Instead maybe give an overview of what you'll learn in this section?  -->

### Limits {#position-continuous-limits}

\index{Axis!limits} \index{Scales!limits}

All scales have limits that specify the values of the aesthetic over which the scale is defined. It's very natural to think about these limits for numeric position scales, as they map directly to the ranges of the axes. By default, the limits are calculated from the range of the data variable, but sometimes you will need to set the limits manually using the `limits` argument to the scale function. Whenever the scale is continuous, as is the case for numeric position scales, this should be a numeric vector of length two. If you only want to set the upper or lower limit, you can set the other value to `NA`. 

Manually setting scale limits is a common task when you need to ensure that scales in different plots are consistent with one another. To illustrate why this is necessary consider this faceted plot:

`r columns(1, aspect_ratio = 1/2, max_width = 1)`
```{r}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() +
  facet_wrap(vars(year))
```

In this plot, ggplot2 has automatically ensured that both facets have the same axis limits, making visual comparison of the two scatter plots easy. However, when creating the plots individually the scale limits in different plots will often be inconsistent:

`r columns(2, 1)`
```{r}
mpg_99 <- mpg %>% filter(year == 1999)
mpg_08 <- mpg %>% filter(year == 2008)

base_99 <- ggplot(mpg_99, aes(displ, hwy)) + geom_point() 
base_08 <- ggplot(mpg_08, aes(displ, hwy)) + geom_point() 

base_99
base_08
```

Each plot makes sense on its own, but visual comparison between the two is difficult due to the inconsistent axis scaling. To ensure consistent axis scaling, we can set the `limits` argument to each scale separately:

```{r}
base_99 + 
  scale_x_continuous(limits = c(1, 7)) +
  scale_y_continuous(limits = c(10, 45))

base_08 + 
  scale_x_continuous(limits = c(1, 7)) +
  scale_y_continuous(limits = c(10, 45))
```

However, this code is a little unwieldy. Because modifying scale limits is such a common task, ggplot2 provides the `lims()` convenience function to simplify the code. Analogous to the `labs()` function used to specify axis labels (Section \@ref(titles)), `lims()` takes name-value pairs as inputs: the argument name is used to specify the aesthetic, and the value is used to specify the scale limits.
\indexf{xlim}  \indexf{ylim} \indexf{lims}

```{r}
base_99 + lims(x = c(1, 7), y = c(10, 45))
base_08 + lims(x = c(1, 7), y = c(10, 45))
```

In the special case where only one axis limit needs to be specified, ggplot2 also provides `xlim()` and `ylim()` helper functions, which can save you a few keystrokes. In practice `lims()` tends to be more useful, because it can be used to set limits for several aesthetics at once. You'll see an example of `lims()` applied to non-position aesthetics in Section \@ref(colour-discrete-limits).

### Zooming in {#zooming-in}

The examples in the previous section expand the scale limits beyond the range spanned by the data. It is also possible to narrow the default scale limits, but care is required: when you truncate the scale limits, some data points will fall outside the boundaries you set, and ggplot2 has to make a decision about what to do with these data points. The default behaviour in ggplot2 is to convert any data values outside the scale limits to `NA`. This means that changing the limits of a scale is not always the same as visually zooming in to a region of the plot. If your goal is to zoom in on part of the plot, it is usually better to use the `xlim` and `ylim` arguments of `coord_cartesian()`:

`r columns(3, 1.5)`
```{r}
base <- ggplot(mpg, aes(drv, hwy)) + 
  geom_hline(yintercept = 28, colour = "red") + 
  geom_boxplot() 

base
base + coord_cartesian(ylim = c(10, 35)) # works as expected
base + ylim(10, 35) # distorts the boxplot 
```

The only difference between the left and middle plots is that the latter is zoomed in. Some of the outlier points are not shown due to the restriction of the range, but the boxplots themselves remain identical. In contrast, in the plot on the right one of the boxplots has changed. When modifying the scale limits, all observations with highway mileage greater than 35 are converted to `NA` before the stat (in this case the boxplot) is computed. Because these "out of bounds" values are no longer available, the end result is that the sample median is shifted downward, which is almost never desirable behaviour. 

<!-- HW: might be worth commenting that this I now regret this behaviour as it's a common source of confusion, but it'd be v hard to change now -->

You can learn more about coordinate systems in Section \@ref(cartesian). To learn more about how "out of bounds" values are handled for continuous and binned scales, see Section \@ref(oob).

### Visual range expansion

If you have eagle eyes, you'll have noticed that the visual range of the axes actually extends a little bit past the numeric limits that I have specified in the various examples. This ensures that the data does not overlap the axes, which is usually (but not always) desirable. You can override the defaults setting the `expand` argument, which expects a numeric vector created by `expansion()`. 

For example, one case where it's usually preferable to remove this space is when using `geom_raster()`, which we can achieve by setting `expand = expansion(0)`: \index{Axis!expansion}

`r columns(2, 1, 1)`
```{r}
base <- ggplot(faithfuld, aes(waiting, eruptions)) + 
  geom_raster(aes(fill = density)) + 
  theme(legend.position = "none") + 
  labs(x = NULL, y = NULL)

base 
base + 
  scale_x_continuous(expand = expansion(0)) + 
  scale_y_continuous(expand = expansion(0)) 
```

Axis expansions are described in terms of an "additive" factor, which specifies a constant space added to outside of the nominal axis limits, and a "multiplicative" one that adds space defined as a proportion of the size of the axis limit. These correspond to the `add` and `mult` arguments to `expansion()`, which can be length one (if the expansion is the same on both sides) or length two (to set different expansions on each side):

`r columns(3, 1, 1)`
```{r}
# additive expansion of three units on both axes
base + 
  scale_x_continuous(expand = expansion(add = 3)) + 
  scale_y_continuous(expand = expansion(add = 3))

# multiplicative expansion of 20% on both axes
base + 
  scale_x_continuous(expand = expansion(mult = .2)) + 
  scale_y_continuous(expand = expansion(mult = .2)) 

# multiplicative expansion of 5% at the lower end of each axes,
# and 20% at the upper end; for the y-axis the expansion is 
# set directly instead of using expansion()
base + 
  scale_x_continuous(expand = expansion(mult = c(.05, .2))) + 
  scale_y_continuous(expand = c(.05, 0, .2, 0))
```

Note the different behaviour in the left and middle plots: the `add` argument is specified on the same scale as the data variable, whereas the `mult` argument is specified relative to the axis range. 


<!-- ### Exercises -->

<!-- 1.  The following code creates two plots of the mpg dataset. Modify the code -->
<!--     so that the legend and axes match, without using faceting! -->

<!--     `r columns(2, 2/3)` -->
<!--     ```{r} -->
<!--     fwd <- subset(mpg, drv == "f") -->
<!--     rwd <- subset(mpg, drv == "r") -->

<!--     ggplot(fwd, aes(displ, hwy, colour = class)) + geom_point() -->
<!--     ggplot(rwd, aes(displ, hwy, colour = class)) + geom_point() -->
<!--     ``` -->

<!-- 1.  What happens if you add two `xlim()` calls to the same plot? Why? -->

<!-- 1.  What does `scale_x_continuous(limits = c(NA, NA))` do? -->

<!-- 1.  What does `expand_limits()` do and how does it work?  Read the source code. -->


### Breaks {#position-continuous-breaks}

Setting the locations of the axis tick marks is a common data visualisation task. In ggplot2, axis tick marks and legend tick marks are both special cases of "scale breaks", and can be modified using the `breaks` argument to the scale function. I'll illustrate this using a toy data set that will reappear in several places throughout this part of the book:

```{r}
toy <- data.frame(
  const = 1, 
  up = 1:4,
  txt = letters[1:4], 
  big = (1:4)*1000,
  log = c(2, 5, 10, 2000)
)
toy
```

To set breaks manually, pass a vector of data values to `breaks`, or set `breaks = NULL` to remove the breaks and the corresponding tick marks entirely. In the plot below, removing the y-axis breaks also removes the corresponding grid lines: 

`r columns(2)`
```{r}
base <- ggplot(toy, aes(big, const)) + 
  geom_point() + 
  labs(x = NULL, y = NULL) +
  scale_y_continuous(breaks = NULL) 

base
```

Alternatively, notice that when the breaks are set manually it moves the major gridlines and the minor gridlines between them:

`r columns(2)`
```{r}
base + scale_x_continuous(breaks = c(1000, 2000, 4000))
base + scale_x_continuous(breaks = c(1000, 1500, 2000, 4000))
```

It is also possible to pass a function to `breaks`. This function should have one argument that specifies the limits of the scale (a numeric vector of length two), and it should return a numeric vector of breaks. You can write your own break function, but in many cases there is no need, thanks to the scales package [@scales]. It provides several tools that are useful for this purpose:

* `scales::breaks_extended()`	creates automatic breaks for numeric axes.
* `scales::breaks_log()`	creates breaks appropriate for log axes.
* `scales::breaks_pretty()`	creates "pretty" breaks for date/times.
* `scales::breaks_width()`	creates equally spaced breaks.

The `breaks_extended()` function is the standard method used in ggplot2, and accordingly the first two plots below are the same. I can alter the desired number of breaks by setting `n = 2`, as illustrated in the third plot. Note that `breaks_extended()` treats `n` as a suggestion rather than a strict constraint. If you need to specify exact breaks it is better to do so manually.  
`r columns(3)`
```{r}
base 
base + scale_x_continuous(breaks = scales::breaks_extended())
base + scale_x_continuous(breaks = scales::breaks_extended(n = 2))
```

Another approach that is sometimes useful is specifying a fixed `width` that defines the spacing between breaks. The `breaks_width()` function is used for this. The first example below shows how to fix the width at a specific value; the second example illustrates the use of the `offset` argument that shifts all the breaks by a specified amount:

`r columns(3)`
```{r}
base + 
  scale_x_continuous(breaks = scales::breaks_width(800))
base + 
  scale_x_continuous(breaks = scales::breaks_width(800, offset = 200))
base + 
  scale_x_continuous(breaks = scales::breaks_width(800, offset = -200))
```

Notice the difference between setting an offset of 200 and -200.

### Minor breaks {#minor-breaks}
\index{Minor breaks}\index{Log!ticks}

You can adjust the minor breaks (the unlabelled faint grid lines that appear between the major grid lines) by supplying a numeric vector of positions to the `minor_breaks` argument. 

Minor breaks are particularly useful for log scales because they give a clear visual indicator that the scale is non-linear. To show them off, I'll first create a vector of minor break values (on the transformed scale), using `%o%` to quickly generate a multiplication table and `as.numeric()` to flatten the table to a vector.

```{r}
mb <- unique(as.numeric(1:10 %o% 10 ^ (0:3)))
mb
```

The following plots illustrate the effect of setting the minor breaks:
`r columns(2, 2/3)`
```{r}
base <- ggplot(toy, aes(log, const)) + 
  geom_point() + 
  labs(x = NULL, y = NULL) +
  scale_y_continuous(breaks = NULL) 

base + scale_x_log10()
base + scale_x_log10(minor_breaks = mb)
```

As with `breaks`, you can also supply a function to `minor_breaks`, such as `scales::minor_breaks_n()` or `scales::minor_breaks_width()` functions that can be helpful in controlling the minor breaks.

### Labels {#position-continuous-labels}

Every break is associated with a label and these can be changed by setting the `labels` argument to the scale function:

`r columns(2, 2/3)`
```{r}
base <- ggplot(toy, aes(big, const)) + 
  geom_point() + 
  labs(x = NULL, y = NULL) +
  scale_y_continuous(breaks = NULL) 

base
base + 
  scale_x_continuous(
    breaks = c(2000, 4000), 
    labels = c("2k", "4k")
  ) 
```

Often you don't need to set the `labels` manually, and can instead specify a labelling function in the same way you can for `breaks`. A function passed to `labels` should accept a numeric vector of breaks as input and return a character vector of labels (the same length as the input). Again, the scales package provides a number of tools that will automatically construct label functions for you. Some of the more useful examples for numeric data include:

* `scales::label_bytes()` formats numbers as kilobytes, megabytes etc.
* `scales::label_comma()` formats numbers as decimals with commas added.
* `scales::label_dollar()` formats numbers as currency.
* `scales::label_ordinal()` formats numbers in rank order: 1st, 2nd, 3rd etc.
* `scales::label_percent()` formats numbers as percentages.
* `scales::label_pvalue()` formats numbers as p-values: <.05, <.01, .34, etc.

A few examples are shown below to illustrate how these functions are used:

`r columns(3)`
```{r breaks-functions}
base <- ggplot(toy, aes(big, const)) + 
  geom_point() + 
  labs(x = NULL, y = NULL) +
  scale_x_continuous(breaks = NULL)

base
base + scale_y_continuous(labels = scales::label_percent())
base + scale_y_continuous(
  labels = scales::label_dollar(prefix = "", suffix = "€")
)
```

You can suppress labels with `labels = NULL`. This will remove the labels from the axis or legend while leaving its other properties unchanged. Notice the difference between setting `breaks = NULL` and `labels = NULL`:

`r columns(2)`
```{r}
base + scale_y_continuous(breaks = NULL)
base + scale_y_continuous(labels = NULL)
```

<!-- ### Exercises -->

<!-- 1.  Recreate the following graphic: -->

<!--     ```{r, echo = FALSE} -->
<!--     ggplot(mpg, aes(displ, hwy)) +  -->
<!--       geom_point() +   -->
<!--       scale_x_continuous("Displacement", labels = scales::unit_format(suffix = "L")) +  -->
<!--       scale_y_continuous(quote(paste("Highway ", (frac(miles, gallon)))))  -->
<!--     ``` -->

<!--     Adjust the y axis label so that the parentheses are the right size. -->

<!-- 1.  List the three different types of object you can supply to the -->
<!--     `breaks` argument. How do `breaks` and `labels` differ? -->

<!-- 1.  What label function allows you to create mathematical expressions? -->
<!--     What label function converts 1 to 1st, 2 to 2nd, and so on? -->


### Transformations {#scale-transformation}

When working with continuous data, the default is to map linearly from the data space onto the aesthetic space. It is possible to override this default using transformations. Every continuous scale takes a `trans` argument, allowing the use of a variety of transformations:
\index{Scales!position} \index{Transformation!scales} \index{Log!scale} \indexf{scale\_x\_log10}

`r columns(2, 1)`
```{r}
# convert from fuel economy to fuel consumption
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  scale_y_continuous(trans = "reciprocal")

# log transform x and y axes
ggplot(diamonds, aes(price, carat)) + 
  geom_bin2d() + 
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10")
```

The transformation is carried out by a "transformer", which describes the transformation, its inverse, and how to draw the labels. You can construct your own transformer using `scales::trans_new()`, but as the plots above illustrate, ggplot2 understands many common transformations supplied by the scales package. The following table lists some of the more common variants: 

| Name                   | Transformer                  | Function $f(x)$         | Inverse $f^{-1}(x)$
|------------------------|------------------------------|-------------------------|------------------------
| `"asn"`                | `scales::asn_trans()`        | $\tanh^{-1}(x)$         | $\tanh(y)$
| `"exp"`                | `scales::exp_trans()`        | $e ^ x$                 | $\log(y)$
| `"identity"`           | `scales::identity_trans()`   | $x$                     | $y$
| `"log"`                | `scales::log_trans()`        | $\log(x)$               | $e ^ y$
| `"log10"`              | `scales::log10_trans()`      | $\log_{10}(x)$          | $10 ^ y$
| `"log2"`               | `scales::log2_trans()`       | $\log_2(x)$             | $2 ^ y$
| `"logit"`              | `scales::logit_trans()`      | $\log(\frac{x}{1 - x})$ | $\frac{1}{1 + e(y)}$
| `"probit"`             | `scales::probit_trans()`     | $\Phi(x)$               | $\Phi^{-1}(y)$
| `"reciprocal"`         | `scales::reciprocal_trans()` | $x^{-1}$                | $y^{-1}$
| `"reverse"`            | `scales::reverse_trans()`    | $-x$                    | $-y$
| `"sqrt"`               | `scales::scale_x_sqrt()`     | $x^{1/2}$               | $y ^ 2$


You can specify the `trans` argument as a string containing the name of the transformation, or by calling the transformer directly. The following are equivalent:

```{r, fig.show='hide'}
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  scale_y_continuous(trans = "reciprocal")

ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  scale_y_continuous(trans = scales::reciprocal_trans())
```

In a few cases ggplot2 simplifies this even further, and provides convenience functions for the most common transformations: `scale_x_log10()`, `scale_x_sqrt()` and `scale_x_reverse()` provide the relevant transformation on the x axis, with similar functions provided for the y axis. Thus, these two plot specifications are also equivalent:

```{r, fig.show="hide"}
ggplot(diamonds, aes(price, carat)) + 
  geom_bin2d() + 
  scale_x_continuous(trans = "log10") +
  scale_y_continuous(trans = "log10")

ggplot(diamonds, aes(price, carat)) + 
  geom_bin2d() + 
  scale_x_log10() +
  scale_y_log10()
```

Note that there is nothing preventing you from performing these transformations manually. For example, instead of using `scale_x_log10()` to transform the scale, you could transform the data instead and plot `log10(x)`. The appearance of the geom will be the same, but the tick labels will be different. Specifically, if you use a transformed scale, the axes will be labelled in the original data space; if you transform the data, the axes will be labelled in the transformed space. As a consequence, these plot specifications are slightly different: 

`r columns(2, 1)`
```{r}
# manual transformation
ggplot(mpg, aes(log10(displ), hwy)) + 
  geom_point()

# transform using scales
ggplot(mpg, aes(displ, hwy)) + 
  geom_point() + 
  scale_x_log10()
```

Regardless of which method you use, the transformation occurs before any statistical summaries. To transform _after_ statistical computation use `coord_trans()`. See Section \@ref(cartesian) for more details on coordinate systems, and Section \@ref(scale-transformation-extras) if you need to transform something other than a numeric position scale.

## Date-time position scales {#date-scales}
\index{Date/times} \index{Data!date/time} \index{Time} \index{Scales!date/time} \indexf{scale\_x\_datetime}

A special case of numeric position arises when an aesthetic is mapped to a date/time type. Examples of date/time types include the base `Date` (for dates) and `POSIXct` (for date-times) classes, as well as the `hms` class for "time of day" values provided by the hms package [@hms]. If your dates are in a different format you will need to convert them using `as.Date()`, `as.POSIXct()` or `hms::as_hms()`. You may also find the lubridate package helpful to manipulate date/time data [@lubridate]. 

Assuming you have appropriately formatted data mapped to the x aesthetic, ggplot2 will use `scale_x_date()` as the default scale for dates and `scale_x_datetime()` as the default scale for date-time data. The corresponding scales for other aesthetics follow the usual naming rules. Date scales behave similarly to other continuous scales, but contain additional arguments that allow you to work in date-friendly units. This section discusses date/time scales for position aesthetics: see Section \@ref(date-colour-scales) for colour and fill aesthetics.

### Breaks 

The `date_breaks` argument allows you to position breaks by date units (years, months, weeks, days, hours, minutes, and seconds). For example, `date_breaks = "2 weeks"` will place a major tick mark every two weeks and `date_breaks = "25 years"` will place them every 25 years:

`r columns(2, 1 / 2)`
```{r date-scale}
date_base <- ggplot(economics, aes(date, psavert)) + 
  geom_line(na.rm = TRUE) +
  labs(x = NULL, y = NULL)

date_base 
date_base + scale_x_date(date_breaks = "25 years")
```

It may be useful to note that internally `date_breaks = "25 years"` is treated as a shortcut for `breaks = scales::breaks_width("25 years")`. The longer form is typically unnecessary, but it can be useful if---as discussed in Section \@ref(position-continuous-breaks)---you wish to specify an `offset`. Suppose the goal is to plot data that span the 20th century, beginning 1 January 1900, and we wish to set breaks in 25 year intervals. Specifying `date_breaks = "25 years"` produces breaks in the following fashion:

```{r}
century20 <- as.Date(c("1900-01-01", "1999-12-31"))
breaks <- scales::breaks_width("25 years")
breaks(century20)
```

<!-- HW: I think this needs a little more motivation -->

Because the range in `century20` starts on 1 January and the breaks increment in whole year values, each of the generated break dates falls on 1 January. We can shift all these breaks so that they fall on 1 February by setting `offset = 31` (since there are thirty one days in January).

### Minor breaks

For date/time scales, you can use the `date_minor_breaks` argument:

`r columns(2, 2/3)`
```{r date-scale-2}
date_base + scale_x_date(
  limits = as.Date(c("2003-01-01", "2003-04-01")),
  date_breaks = "1 month"
)

date_base + scale_x_date(
  limits = as.Date(c("2003-01-01", "2003-04-01")),
  date_breaks = "1 month",
  date_minor_breaks = "1 week"
)
```

<!-- HW: maybe date_breaks = "1 year" would make the minor breaks easier to see? Or maybe labels = NULL? -->

Note that in the first plot, the minor breaks are spaced evenly between the monthly major breaks. In the second plot, the major and minor beaks follow slightly different patterns: the minor breaks are always spaced 7 days apart but the major breaks are 1 month apart. Because the months vary in length, this leads to slightly uneven spacing.


### Labels {#date-labels}

Date scales contain a `labels` argument that behaves similarly to the corresponding argument for numeric scales, but is often more convenient to use the `date_labels` argument. It controls the display of the labels using the same formatting strings as in `strptime()` and `format()`. To display dates like 14/10/1979, for example, you would use the  string `"%d/%m/%Y"`: in this expression `%d` produces a numeric day of month, `%m` produces a numeric month, and `%Y` produces a four digit year. The table below provides a list of formatting strings:

| String        | Meaning
|:--------------|:-----------------------------------------
| `%S`          | second (00-59)
| `%M`          | minute (00-59)
| `%l`          | hour, in 12-hour clock (1-12)
| `%I`          | hour, in 12-hour clock (01-12)
| `%p`          | am/pm
| `%H`          | hour, in 24-hour clock (00-23)
| `%a`          | day of week, abbreviated (Mon-Sun)
| `%A`          | day of week, full (Monday-Sunday)
| `%e`          | day of month (1-31)
| `%d`          | day of month (01-31)
| `%m`          | month, numeric (01-12)
| `%b`          | month, abbreviated (Jan-Dec)
| `%B`          | month, full (January-December)
| `%y`          | year, without century (00-99)
| `%Y`          | year, with century (0000-9999)


One useful scenario for date label formatting is when there's insufficient room to specify a four digit year. Using `%y` ensures that only the last two digits are displayed:

`r columns(2, 1 / 2)`
```{r date-scale-3}
base <- ggplot(economics, aes(date, psavert)) + 
  geom_line(na.rm = TRUE) +
  labs(x = NULL, y = NULL)

base + scale_x_date(date_breaks = "5 years")
base + scale_x_date(date_breaks = "5 years", date_labels = "%y")
```

It can be useful to include the line break character `\n` in a formatting string, particularly when full-length month names are included:

```{r date-scale-4}
lim <- as.Date(c("2004-01-01", "2005-01-01"))

base + scale_x_date(limits = lim, date_labels = "%b %y")
base + scale_x_date(limits = lim, date_labels = "%B\n%Y")
```

In these examples I have specified the labels manually via the `date_labels` argument. An alternative approach is to pass a labelling function to the `labels` argument, in the same way I described in Section \@ref(position-continuous-labels). The scales package provides two convenient functions that will generate date labellers for you:

<!-- HW: weird to have bulleted list with only two elements. Maybe describe label_date() as what date_labels does for you (like with breaks), and then have a paragraph about label_date_short()? -->

* `label_date()` is what `date_labels` does for you behind the scenes, so you
  rarely need to call it directly.

* `label_date_short()` automatically constructs short labels that are 
  sufficient to uniquely identify the dates: 

    ```{r date-scale-5}
    base + scale_x_date(labels = scales::label_date("%b %y"))
    base + 
      scale_x_date(
        limits = lim, 
        labels = scales::label_date_short()
      )
    ```


## Discrete position scales

It is also possible to map discrete variables to position scales, with the default scales being `scale_x_discrete()` and `scale_y_discrete()`. For example, the following two plot specifications are equivalent

```{r default-scales-discrete, fig.show = "hide"}
ggplot(mpg, aes(x = hwy, y = class)) + 
  geom_point()

ggplot(mpg, aes(x = hwy, y = class)) + 
  geom_point() + 
  scale_x_continuous() +
  scale_y_discrete()
```

Internally, ggplot2 handles discrete scales by mapping each category to an integer value and then drawing the geom at the corresponding coordinate location. To illustrate this, we can add a custom annotation (see Section \@ref(custom-annotations)) to the plot:

`r columns(1)`
```{r}
ggplot(mpg, aes(x = hwy, y = class)) + 
  geom_point() +
  annotate("text", x = 5, y = 1:7, label = 1:7)
```

Mapping each category to an integer value is useful because it means that other width quantities can be specified as a proportion of the category range. For instance, in the preceding plot, we could specify a vertical jitter for each point spanning half the width of the implied category bin:

`r columns(1)`
```{r}
ggplot(mpg, aes(x = hwy, y = class)) + 
  geom_jitter(width = 0, height = .25) +
  annotate("text", x = 5, y = 1:7, label = 1:7)
```

The same mechanism underpins the widths of bars and boxplots. Because each category has width 1 in a discrete scale, setting `width = .4` when using `geom_boxplot()` ensures that the box occupies 40% of the width allocated to the category:

`r columns(2)`
```{r}
ggplot(mpg, aes(x = drv, y = hwy)) + geom_boxplot()
ggplot(mpg, aes(x = drv, y = hwy)) + geom_boxplot(width = .4)
```


### Limits, breaks, and labels {#scale-labels}
\index{Axis!labels} \index{Legend!keys}

The limits, breaks. and labels for a discrete position scale can be set using the `limits`, `breaks`, and `labels` arguments. For the most part these behave identically to the corresponding arguments for numeric scales (Section \@ref(numeric-position-scales)), though there are some differences. For example, the limits of a discrete scale are not defined in terms of endpoints, but instead correspond to the set of allowable values for that variable. Accordingly, ggplot2 expects that the `limits` of a discrete scale should be a character vector that enumerates all possible values in the order they should appear:

`r columns(3)`
```{r}
base <- ggplot(toy, aes(const, txt)) + 
  geom_label(aes(label = txt)) +
  scale_x_continuous(breaks = NULL) +
  labs(x = NULL, y = NULL)

base 
base + scale_y_discrete(limits = c("a", "b", "c", "d", "e"))
base + scale_y_discrete(limits = c("d", "c", "a", "b"))
```

The `breaks` argument is largely unchanged, enumerating a set of values to be displayed on the axis labels. The `labels` argument for discrete scales has some additional functionality: you also have the option of using a named vector to set the labels associated with particular values. This allows you to change some labels and not others, without altering the ordering or the breaks:

`r columns(2, 2/3)`
```{r}
base + scale_y_discrete(breaks = c("b", "c"))
base + scale_y_discrete(labels = c(c = "carrot", b = "banana")) 
```

As with other scales, discrete position scales allow you to pass a function to the `labels` argument. The `scales::label_wrap()` function can be particularly valuable for categorical data, as it allows you to wrap long strings across multiple lines. 

### Label positions {#guide-axis}

When plotting categorical data it is often necessary to move the axis labels in some way to prevent them from overlapping. You can do this easily with the help of the `guide` argument and the `guide_axis()` function. For example, to "dodge" the position of the labels, you can set `guide_axis(n.dodge = 3)`:

`r columns(2, 2/3)`
```{r}
base <- ggplot(mpg, aes(manufacturer, hwy)) + geom_boxplot() 

base
base + scale_x_discrete(guide = guide_axis(n.dodge = 3))
```

An alternative way to accomplish the same task is with the `guides()` function 

<!-- HW: again feels like we should consistently use `guides()` I think-->

The `guides()` helper works in a similar way to the `labs()` helper function described in Section \@ref(titles). Both take the name of different aesthetics (e.g., color, x, fill) as arguments and allow you to specify your own value. Where `labs()` provides a shorthand way to specify the `name` argument to one or more scales, the `guides()` function allows you to specify `guide` arguments to one or more scales. In the same way that `labs(x = "an axis label")` specifies the name associated with the x-axis scale, a command such as `guides(x = guide_axis())` can be used to specify its associated guide. For example, if you want to rotate the axis labels, either of the following will work:

`r columns(2, 2/3)`
```{r}
base + scale_x_discrete(guide = guide_axis(angle = 90))
base + guides(x = guide_axis(angle = 90))
```

To learn more about guide functions see Section \@ref(scale-guide).

## Binned position scales {#binned-position}

A variation on discrete position scales are binned scales, where a continuous variable is sliced into multiple bins and the discretised variable is plotted. For example, if we want to modify the plot above to show the number of observations at each location, we could use `geom_count()` instead of `geom_point()` so that the size of the dots scales with the number of observations. As the left plot below illustrates, this is an improvement but is still rather cluttered. To improve this, the plot on the right uses `scale_x_binned()` to cut the `hwy` values into 10 bins before passing them to the geom:

`r columns(2, 2/3)`
```{r}
base <- ggplot(mpg, aes(hwy, class)) + geom_count()

base 
base + scale_x_binned(n.breaks = 10)
```

<!-- HW: this feels a bit brief. Maybe Thomas will have some ideas for more content? -->

