\input{header.tex}

\setchapterpreamble[u]{% 
\dictum[Anonymous]{Forecasting is the art of saying 
what is going to happen and then to explain 
why it didn’t.}} 

\chapter{Scales}

\section{Introduction}\label{sec:introduction}

Scales control the mapping between data space and aesthetic space.  They convert your data values into aesthetic attributes that can be a perceived: colour, shape, size, etc.  Each type of aesthetic attribute has a default scale, and may have other scales that provide different types of mappings.  For example, the default colour scale uses equally space hues, but other scales allow you to generate a gradient between two or three different colours.  This section describes the basic operation of a scale, the details of all the different scales and instructions on how to make your own.

\section{Legend and axis labels}\label{sec:legend_and_axis_labels} % (fold)

> Is it possible to change the legend title in ggplot that is generated
> by a colour (other than by changing the name of the variable)?  For
> example, this produces a legend title, "tip/total_bill".  How can that
> title be changed?
> 
> ggpoint(ggplot(tips,aesthetics=list(y=tip,x=total_bill,colour=tip/total_bill)) )

You need to add a colour scale, and specify the name to appear on the legend (by default it just uses the name of the variable):

p <- ggpoint(ggplot(tips,aesthetics=list(y=tip,x=total_bill,colour=tip/total_bill)) )
sccolour(p, "Tip rate")
sccolour(p, "This is a very long label\nsplit over two lines")

you can also use expressions to create mathematical annotations:

sccolour(p, expression(beta * x^2))

% section legend_and_axis_labels (end)


\section{Aesthetic attributes}\label{sec:aesthetic_attributes}

The following aesthetic attributes are available in {\tt ggplot}.  This list is not exhaustive, other authors may add grob functions that use other mappings.  

Any aesthetic that a grob does not understand will be silently ignored.  If a scale is not doing what you expect, make sure to check the you don't have any spelling mistakes.

Position scales
Other scales
 * categorical
 * continuous

\begin{itemize}
	\item colour
	\item fill
	\item group
	\item id
	\item linetype
	\item shape
	\item size
	\item dimension: height, width
	\item position: x, y, z
\end{itemize}

There are also a couple of aesthetic attributes that can't be perceived directly.  The most important of these is the {\tt id} aesthetic, which divides the the data set into discrete components.   This is used in line and path plots to separate the data for different lines, and in the groups grob to divide the different groups. 

Scales may take multiple inputs and return multiple outputs.

\section{Using scales}\label{sec:using_scales}

Scales are added by default whenever you use {\tt qplot} or {\tt ggplot}.  If you want to modify them, you need to manually add a scale to the plot.  This will automatically override any defaults.

% decumar<<< 
% interweave({
% quakes$depthc <- chop(quakes$depth)
% p <- qplot(long, lat, . ~ depthc, data=quakes)
% str(p$scales)
% p <- pscontinuous(p, "x", breaks=c(5,10))
% str(p$scales)
% })
% |||
% >>>

All position scales start with {\tt sc}, all others with {\tt sc}.

\section{How to write your own scales}\label{sec:how_to_write_your_own_scales}


\input{footer.tex}
