---
title: specifications
output: bookdown::html_chapter
bibliography: references.bib
---


```{r specifications, echo=FALSE}
knitr::opts_chunk$set(eval = FALSE)
```

# Aesthetic specifications {#cha:specifications}

This appendix summarises the various formats that **grid** drawing functions take.  Most of this information is available scattered throughout the R documentation.  This appendix brings it all together in one place. \index{Aesthetics!specifications}

## Colour {#sec:colour_spec}

Colours can be specified with: \index{Colour!specifying}

* A __name__, e.g., `"red"`. The colours are displayed in Figure \ref{fig:colours}, and can be listed in more detail with `colours()`. The Stowers Institute provides a nice printable pdf that lists all colours:  <http://research.stowers-institute.org/efg/R/Color/Chart/>.
* An __rgb specification__, with a string of the form `"#RRGGBB"` where each of the pairs `RR`, `GG`, `BB` consists of two hexadecimal digits giving a value in the range `00` to `FF`.  Colors can be made partially transparent with `alpha`, e.g., `alpha = 0.5`.
* An __NA__, for a completely transparent colour.

The functions `rgb()`, `hsv()`, `hcl()` can be used to create colours specified in different colour spaces.

<!--
% The `diverge_hcl()`, `sequential_hcl()`, `rainbow_hcl()`, and `heat_hcl()` functions from the **vcd** package provide other ways of generate colours palettes based on perceptually sound principles.
-->

## Line type {#sec:line-type-spec}

Line types can be specified with: \index{Line type!specifying}

* An __integer__ or __name__: 0=blank, 1=solid, 2=dashed, 3=dotted, 4=dotdash, 5=longdash, 6=twodash), illustrated in Figure \ref{fig:linetype}.
* The lengths of on/off stretches of line. This is done with a string of an even number (up to eight) of hexadecimal digits which give the lengths in consecutive positions in the string. For example, the string `"33"` specifies three units on followed by three off and `"3313"` specifies three units on followed by three off followed by one on and finally three off. 
  
The five standard dash-dot line types described above correspond to 44, 13, 134, 73 and 2262.

Note that `NA` is not a valid value for `lty`.

## Shape {#sec:shape-spec}

Shapes take four types of values: \index{Shape!specifying}

* An __integer__ in $[0, 25]$, illustrated in Figure \ref{fig:shape}.
* A __single character__, to use that character as a plotting symbol.
* A `.` to draw the smallest rectangle that is visible (i.e., about one pixel).
* An `NA`, to draw nothing.

While all symbols have a foreground colour, symbols 19--25 also take a background colour (fill). 

## Size {#sec:size}

Throughout `ggplot`, for text height, point size and line width, size is specified in millimetres. \index{Size!specifying}

```{r justification, echo = FALSE}
# This code should recreate the figures, but I don't think we can do fancy subfigures with knitr :(
library(grid)
xquiet <- scale_x_continuous("", breaks = NA)
yquiet <- scale_y_continuous("", breaks = NA)
fill <- theme(
  plot.margin = unit(rep(0, 4), "cm"), 
  axis.text.x = element_blank(),
  axis.text.y = element_blank(),
  axis.title.x = element_blank(),
  axis.title.y = element_blank(),
  axis.ticks.length = unit(0, "cm"),
  axis.ticks.margin = unit(0, "cm"),
  panel.grid.major = element_blank(),
  panel.grid.minor = element_blank(),
  plot.background = element_rect(fill = "grey90", colour = NA)
)

# Shapes ---------------------------------------------------------
shapes <- data.frame(
  shape = c(0:19, 22, 21, 24, 23, 20),
  x = 0:24 %/% 5,
  y = -(0:24 %% 5)
)
qplot(x, y, data=shapes, shape=shape, size=I(5), fill=I("blue")) +
  scale_shape_identity() + xlim(0, 4.4) + 
  geom_text(aes(x = x + 0.2, label=shape), hjust=0) + fill

# Line types -----------------------------------------------------------------
lty <- c("blank", "solid", "dashed", "dotted", "dotdash", 
         "longdash","twodash")
linetypes <- data.frame(
  y = seq_along(lty),
  lty = lty
) 
qplot(0, y, data=linetypes, xend = 5, yend=y, geom="segment", linetype=lty) +
  scale_linetype_identity() + 
  geom_text(aes(x = 0, y = y + 0.2, label = lty), hjust = 0) + 
  fill


# Colour ---------------------------------------------------------------------
source("colour-wheel.R")
qplot(x, y, data=hcl, colour=colour) + 
  scale_colour_identity() + 
  coord_equal() + 
  fill
```


\begin{figure}[htbp]
  \centering
  \subfigure[All named colours in Luv space]{
    \label{fig:colours}
    \includegraphics[width=0.45\linewidth]{diagrams/spec-colour}
  }
  \subfigure[Built-in line types]{
    \label{fig:linetype}
    \includegraphics[width=0.45\linewidth]{diagrams/spec-linetype}
  }
  \subfigure[R plotting symbols.  Colour is black, and fill is blue.  Symbol 25 (not shown) is symbol 24 rotated 180 degrees.]{
    \label{fig:shape}
    \includegraphics[width=0.45\linewidth]{diagrams/spec-shape}  
  }
  \subfigure[Horizontal and vertical justification settings.]{
    \label{fig:justification}
    \includegraphics[width=0.45\linewidth]{diagrams/spec-justification}
  }
  \caption{Examples illustrating different aesthetic settings.}
\end{figure}

<!--
% \section{Fonts}
% \label{sec:fonts}
% 
% postscriptFonts, pdfFonts, quartzFonts
% 
% Find R news article
% 
% \begin{itemize}
%   \item \code{face}
%   \item \code{family}
%   \item \code{lineheight}
%   \item \code{fontsize}
% \end{itemize}
-->
